# Library installation history and updates

## R 4.5.2 (Current Setup - March 2026)
- **Library Paths**: Created centralized library directories in `env/R_4.5.2/` (`ArchR_libs` and `giotto_libs`) to keep the environment clean and compatible with the new R version.
- **Dependency Management**: Updated `devtools` and `BiocManager` installation to use these specific local library paths.
- **Giotto Updates**: Switched from version-tagged GitHub installs (e.g., `@R4.4.0`) to `ref="master"` to ensure compatibility with R 4.5.2's updated base packages (like `Matrix` and `terra`).
- **Fixes**: 
    - Re-ran explicit `remotes::install_github("drieslab/Giotto")` to ensure the main package was correctly linked after its sub-components (`GiottoUtils`, `GiottoClass`, `GiottoVisuals`) were verified.
    - Fixed `terra` installation by explicitly loading system modules (`gdal/3.8.4`, `geos/3.12.1`, `proj/9.3.1`) before running the install to resolve the missing `gdal-config` error.
    - Specifically installed `terra 1.9-1` into the local `giotto_libs` path using `--no-lock` to satisfy `pak::pkg_install("drieslab/Giotto")` requirements for the newest version.
- **Automation**: Updated `.project_modules` to automatically export `R_LIBS_USER` pointing to these new directories whenever the workspace is loaded.

## R 4.4.0 (Previous Setup)
- Libraries were installed in `/projectnb/paxlab/presh/giotto_libs` and specific paths for ArchR that were versioned for 4.4.0.
- Giotto components were installed using specific branch tags (`@R4.4.0`) to maintain stability within that environment.
- Some dependencies like `Matrix` were pinned to `1.6-5` to avoid breaking changes in the older R version.

### R 4.4.0 Installation Commands (for reference)
```r
# ArchR 4.4.0
.libPaths('/projectnb/paxlab/presh/env/R_4.4.0/ArchR_libs')
devtools::install_github("GreenleafLab/ArchR", ref="master")

# Giotto 4.4.0
.libPaths('/projectnb/paxlab/presh/env/R_4.4.0/giotto_libs')

if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")


devtools::install_github("drieslab/GiottoUtils", ref="master", force = T)

# If terra gives error run the 2 lines below 
# unlink("/projectnb/paxlab/presh/env/R_4.5.2/giotto_libs/00LOCK-terra", recursive = TRUE)
# remotes::install_github("rspatial/terra")
packageVersion('terra') # need 1.8.7 
# devtools::install_version("Matrix", version = "1.6-5")
devtools::install_github("drieslab/GiottoClass", ref="master", force = T)
devtools::install_github("drieslab/GiottoVisuals", ref="master", force = T)
devtools::install_github("drieslab/Giotto", ref="master", force = T)


library(Giotto)

# Ensure Giotto can access the python env
set_giotto_python_path(python_path = '/usr3/graduate/preshita/.conda/envs/giotto_env/bin/python', verbose = NULL, initialize = TRUE)

# Installing python env 
Giotto::installGiottoEnvironment()

reticulate::py_install(
    pip = TRUE,
    envname = 'giotto_env',
    packages = c(
        "scanpy"
    )
)

# install another environment with py 3.8 for cellpose
# reticulate::conda_create(envname = "giotto_cellpose",
#                          python_version = 3.8)
# this command not working so doing through conda

# do in linux
# conda create -n giotto_cellpose python=3.8

# Restart R session 
reticulate::use_condaenv('giotto_cellpose')
reticulate::py_install(
  pip = TRUE,
  envname = 'giotto_cellpose',
  packages = c(
    "pandas",
    "networkx",
    "python-igraph",
    "leidenalg",
    "scikit-learn",
    "cellpose",
    "smfishhmrf",
    'tifffile',
    'scikit-image'
  )
)

```

---
title: "Library installation for R-4.5.2"
author: "Preshita Date"
date: "March 2026"
---

This notebook is for making ArchR and Giotto libraries at separate paths for R 4.5.2 

# Creating ArchR library

First load these modules into VS code or interactive session - 
module load gdal/3.8.4
module load geos/3.12.1
module load proj/9.3.1
module load sqlite3/3.44.2
module load openblas/0.3.23 
module load gcc/12.2.0
module load glpk/5.0
module load R/4.5.2

module load R 
R


```{r}
.libPaths('/projectnb/paxlab/presh/env/R_4.5.2/ArchR_libs')
.libPaths()

if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
devtools::install_github("GreenleafLab/ArchR", ref="master", repos = BiocManager::repositories())
```

```{r}
library(ArchR)

# RcppArmadillo gave me issues, use this to install!
install.packages("RcppArmadillo", type = "source",
  configure.args = c(
    "--with-lapack",
    "--with-blas"
  ))
```


# Creating Giotto library

```{r}
.libPaths('/projectnb/paxlab/presh/env/R_4.5.2/giotto_libs')
.libPaths()

if(!"pak" %in% installed.packages()) {
  install.packages("pak")
}

pak::pkg_install("drieslab/Giotto")
remotes::install_github("rspatial/terra@1.9-1")


```




# To load libraries in future sessions, just load the modules and then do -

```{r}
.libPaths(c("/projectnb/paxlab/presh/env/R_4.5.2/ArchR_libs", "/projectnb/paxlab/presh/env/R_4.5.2/giotto_libs", .libPaths()))
library(ArchR)
library(Giotto)

```


Also have to set up a conda env for zellkonverter for running Philipp's code

```{bash}
module load miniconda/25.3.1
# Create the environment at the specific prefix
conda create --prefix /projectnb/paxlab/presh/env/conda_env/cellwalkr_env python=3.13 -y

# Activate the environment using its path
conda activate /projectnb/paxlab/presh/env/conda_env/cellwalkr_env

# Install the anndata package (required by zellkonverter)
pip install anndata h5py
```










