---
title: "MOSAICField Alignment Validation: ATAC Spots to Xenium Cells Mapping"
date: "`r Sys.Date()`"
---

## Setup Instructions

Before running this notebook, set up the environment:

```bash
conda activate /projectnb/paxlab/presh/env/conda_env/giotto_env311
export LD_LIBRARY_PATH=/projectnb/paxlab/presh/env/conda_env/giotto_env311/lib:$LD_LIBRARY_PATH
export LD_PRELOAD=/projectnb/paxlab/presh/env/conda_env/giotto_env311/lib/libstdc++.so.6
R --vanilla
```

## Overview

This notebook validates MOSAICField alignment outputs and maps spatial ATAC spots to Xenium single-cell locations. 
The workflow includes:
1. **Debugging**: Validate alignment quality and spatial consistency
2. **Polygon Creation**: Create 10µm square polygons for ATAC spots
3. **Mapping**: Calculate Xenium cell to ATAC spot overlap
4. **Analysis**: Generate density statistics and visualization 

## Section 1: Environment Initialization

```{r setup, message=FALSE, warning=FALSE}
# Configure Python environment
Sys.setenv(RETICULATE_PYTHON = "/projectnb/paxlab/presh/env/conda_env/giotto_env311/bin/python")

# Load libraries
library(reticulate)
library(Giotto)
library(data.table)
library(ggplot2)
library(arrow)
library(sf)
library(dplyr)

# Verify Python and packages
py_config()
tryCatch(reticulate::import("scanpy"), error = function(e) warning("scanpy not found"))

# Configure R library paths for Giotto
.libPaths(c('/projectnb/paxlab/presh/env/R_4.5.2/giotto_libs', .libPaths()))

# Initialize Giotto Python backend
GiottoClass::set_giotto_python_path(
  python_path = "/projectnb/paxlab/presh/env/conda_env/giotto_env311/bin/python",
  initialize = TRUE
)

# Set paths
input.path <- '/projectnb/paxlab/presh/projects/spatial_atac/analysis/src/alignment/mosaicfield_outputs'
output.path <- '/projectnb/paxlab/presh/projects/spatial_atac/analysis/src/alignment'

cat("[SETUP] All libraries loaded and paths configured\n")
```

## Section 2: Debugging - Validate Alignment Quality

Before mapping, we validate that MOSAICField correctly aligned the ATAC and Xenium data.

```{r debug-load-data, message=FALSE}
cat("\n=== STEP 1: Loading Aligned Objects ===\n")

# Load ATAC object
cat("Loading ATAC Giotto object...\n")
g_atac <- anndataToGiotto(
  anndata_path = file.path(input.path, "atac_mosaicfield_outputs.h5ad"),
  python_path = "/projectnb/paxlab/presh/env/conda_env/giotto_env311/bin/python"
)
cat("✓ ATAC object loaded\n")

# Load Xenium object
cat("Loading Xenium Giotto object...\n")
g_xenium <- anndataToGiotto(
  anndata_path = file.path(input.path, "xenium_mosaicfield_outputs.h5ad"),
  python_path = "/projectnb/paxlab/presh/env/conda_env/giotto_env311/bin/python"
)
cat("✓ Xenium object loaded\n")

# Extract metadata
atac_meta <- pDataDT(g_atac)
xenium_meta <- pDataDT(g_xenium)

cat("\nObject Dimensions:\n")
cat("  ATAC spots:", nrow(atac_meta), "\n")
cat("  Xenium cells:", nrow(xenium_meta), "\n")
```

```{r debug-validate-coordinates}
cat("\n=== STEP 2: Validate Spatial Coordinates ===\n")

# Check for coordinate columns
atac_has_coords <- all(c("x", "y") %in% colnames(atac_meta))
xenium_has_coords <- all(c("x", "y") %in% colnames(xenium_meta))

cat("ATAC has x,y coordinates:", atac_has_coords, "\n")
cat("Xenium has x,y coordinates:", xenium_has_coords, "\n")

# Print coordinate ranges
if (atac_has_coords) {
  cat("\nATAC Coordinate Range:\n")
  print(data.frame(
    dimension = c("X", "Y"),
    min = c(min(atac_meta$x), min(atac_meta$y)),
    max = c(max(atac_meta$x), max(atac_meta$y)),
    span = c(max(atac_meta$x) - min(atac_meta$x), 
             max(atac_meta$y) - min(atac_meta$y))
  ))
}

if (xenium_has_coords) {
  cat("\nXenium Coordinate Range:\n")
  print(data.frame(
    dimension = c("X", "Y"),
    min = c(min(xenium_meta$x), min(xenium_meta$y)),
    max = c(max(xenium_meta$x), max(xenium_meta$y)),
    span = c(max(xenium_meta$x) - min(xenium_meta$x), 
             max(xenium_meta$y) - min(xenium_meta$y))
  ))
}
```

```{r debug-alignment-quality, message=FALSE}
cat("\n=== STEP 3: Alignment Quality Assessment ===\n")

if (atac_has_coords && xenium_has_coords) {
  # Calculate spatial overlap metrics
  atac_area <- (max(atac_meta$x) - min(atac_meta$x)) * (max(atac_meta$y) - min(atac_meta$y))
  xenium_area <- (max(xenium_meta$x) - min(xenium_meta$x)) * (max(xenium_meta$y) - min(xenium_meta$y))
  
  cat("Bounding Box Areas:\n")
  cat("  ATAC area:", formatC(atac_area, format="f", digits=0), "µm²\n")
  cat("  Xenium area:", formatC(xenium_area, format="f", digits=0), "µm²\n")
  cat("  Area ratio (xenium/atac):", round(xenium_area/atac_area, 3), "\n")
  
  # Assess alignment
  area_ratio <- xenium_area / atac_area
  if (0.5 < area_ratio && area_ratio < 2.0) {
    cat("✓ GOOD: Objects have similar spatial extents (well-aligned)\n")
    alignment_check <- "PASS"
  } else if (0.1 < area_ratio && area_ratio < 10) {
    cat("⚠ CAUTION: Objects have different spatial extents (possible misalignment)\n")
    alignment_check <- "CAUTION"
  } else {
    cat("✗ FAIL: Objects have very different scales (likely misaligned)\n")
    alignment_check <- "FAIL"
  }
  
  cat("\nAlignment Status:", alignment_check, "\n")
}
```

```{r debug-plots, fig.width=12, fig.height=12, message=FALSE}
cat("\n=== STEP 4: Diagnostic Overlay Plots ===\n")

if (atac_has_coords && xenium_has_coords) {
  
  # Plot 1: Full overlay of ATAC and Xenium
  p1 <- ggplot() +
    geom_point(data = atac_meta, aes(x = x, y = y, color = "ATAC Spots"), 
               size = 2, alpha = 0.6, shape = 15) +
    geom_point(data = xenium_meta, aes(x = x, y = y, color = "Xenium Cells"), 
               size = 1, alpha = 0.3, shape = 16) +
    scale_color_manual(values = c("ATAC Spots" = "red", "Xenium Cells" = "blue")) +
    coord_fixed() +
    theme_minimal() +
    labs(
      title = "Overlay: ATAC Spots (red) vs Xenium Cells (blue)",
      subtitle = paste("ATAC:", nrow(atac_meta), "spots | Xenium:", nrow(xenium_meta), "cells"),
      x = "X Coordinate (µm)",
      y = "Y Coordinate (µm)",
      color = "Data Type"
    ) +
    theme(legend.position = "right", plot.title = element_text(face = "bold"))
  
  print(p1)
  ggsave(file.path(output.path, "01_debug_overlay_all.png"), p1, width = 10, height = 10, dpi = 150)
  cat("✓ Saved: 01_debug_overlay_all.png\n")
  
  # Plot 2: ATAC spots only
  p2 <- ggplot(atac_meta, aes(x = x, y = y)) +
    geom_point(color = "red", size = 2.5, alpha = 0.6, shape = 15) +
    coord_fixed() +
    theme_minimal() +
    labs(
      title = paste("ATAC Spots Distribution (n =", nrow(atac_meta), ")"),
      x = "X Coordinate (µm)",
      y = "Y Coordinate (µm)"
    ) +
    theme(plot.title = element_text(face = "bold"))
  
  print(p2)
  ggsave(file.path(output.path, "02_debug_atac_spots.png"), p2, width = 8, height = 8, dpi = 150)
  cat("✓ Saved: 02_debug_atac_spots.png\n")
  
  # Plot 3: Xenium cells only
  p3 <- ggplot(xenium_meta, aes(x = x, y = y)) +
    geom_point(color = "blue", size = 1, alpha = 0.5, shape = 16) +
    coord_fixed() +
    theme_minimal() +
    labs(
      title = paste("Xenium Cells Distribution (n =", nrow(xenium_meta), ")"),
      x = "X Coordinate (µm)",
      y = "Y Coordinate (µm)"
    ) +
    theme(plot.title = element_text(face = "bold"))
  
  print(p3)
  ggsave(file.path(output.path, "03_debug_xenium_cells.png"), p3, width = 8, height = 8, dpi = 150)
  cat("✓ Saved: 03_debug_xenium_cells.png\n")
  
  cat("\nDiagnostic plots saved to:", output.path, "\n")
}
```

## Section 3: ATAC-to-Xenium Mapping - Polygon-Based Alignment
```{r mapping-step1-create-atac-polygons, message=FALSE}
cat("\n=== MAPPING STEP 1: Create 10µm Square Polygons from ATAC Centroids ===\n")

# Get ATAC metadata (already loaded in debug section)
atac_meta <- pDataDT(g_atac)

# Create 4 vertices per ATAC spot to define 10µm squares
# dist = 5µm = half-width of 10µm edge
dist <- 5

# Build polygon data for all ATAC spots
poly_list <- lapply(seq_len(nrow(atac_meta)), function(i) {
  spot_id <- atac_meta$cell_ID[i]
  x_center <- atac_meta$x[i]
  y_center <- atac_meta$y[i]
  
  data.table(
    x = c(x_center - dist, x_center + dist, x_center + dist, x_center - dist),
    y = c(y_center - dist, y_center - dist, y_center + dist, y_center + dist),
    poly_ID = spot_id
  )
})

poly_dt <- rbindlist(poly_list)

cat("Created polygon vertices for", nrow(atac_meta), "ATAC spots\n")
cat("Total vertices:", nrow(poly_dt), "\n")

# Create Giotto polygon object
tryCatch({
  atac_polygons <- createGiottoPolygonsFromDfr(
    segmdfr = as.data.frame(poly_dt),
    name = "atac_bins",
    calc_centroids = TRUE,
    verbose = FALSE
  )
  cat("✓ ATAC polygon object created\n")
}, error = function(e) {
  cat("✗ Error creating polygons:", e$message, "\n")
  stop(e)
})
```

```{r mapping-step2-add-xenium-boundaries, message=FALSE}
cat("\n=== MAPPING STEP 2: Load Xenium Cell Boundaries ===\n")

# Load Xenium cell boundaries from parquet file
cell_bounds_path <- "/projectnb/paxlab/DATA/DriesSpatial/Xenium/output-XETG00253__0048833__488B__20241217__182301/cell_boundaries.parquet"

tryCatch({
  if (!file.exists(cell_bounds_path)) {
    stop("Cell boundaries parquet file not found at:", cell_bounds_path)
  }
  
  cell_bounds_df <- read_parquet(cell_bounds_path)
  cell_bounds_dt <- as.data.table(cell_bounds_df)
  
  cat("✓ Loaded cell boundaries\n")
  cat("  Dimensions:", nrow(cell_bounds_dt), "rows,", ncol(cell_bounds_dt), "columns\n")
  cat("  Columns:", paste(colnames(cell_bounds_dt), collapse=", "), "\n")
  
  # Create sf polygons from boundaries
  cell_polys_sf <- cell_bounds_dt %>%
    group_by(poly_ID) %>%
    summarise(
      geometry = st_polygon(list(as.matrix(cur_data()[, .(x, y)]))),
      .groups = "drop"
    ) %>%
    st_as_sf()
  
  cat("  Created", nrow(cell_polys_sf), "polygon objects from boundaries\n")
  
  # Convert to Giotto polygon format
  cell_polygons <- as.polygons(cell_polys_sf, name = "cell")
  
  # Add to Xenium object
  g_xenium <- addGiottoPolygons(
    gobject = g_xenium,
    gpolygons = list(cell = cell_polygons)
  )
  
  cat("✓ Cell boundaries added to Xenium object\n")
  
}, error = function(e) {
  cat("⚠ Warning: Could not load cell boundaries:", e$message, "\n")
  cat("  Continuing without cell boundaries (mapping will use cell centroids)\n")
})
```

```{r mapping-step3-add-atac-polygons, message=FALSE}
cat("\n=== MAPPING STEP 3: Add ATAC Polygons to Xenium Object ===\n")

tryCatch({
  g_xenium <- addGiottoPolygons(
    gobject = g_xenium,
    gpolygons = list(atac_bins = atac_polygons)
  )
  
  cat("✓ ATAC polygons added to Xenium object\n")
  
  # Verify polygons were added
  poly_info <- getPolygonInfo(g_xenium)
  cat("  Available polygons:\n")
  print(poly_info)
  
}, error = function(e) {
  cat("✗ Error adding ATAC polygons:", e$message, "\n")
  stop(e)
})
```

```{r mapping-step4-calculate-overlap, message=FALSE}
cat("\n=== MAPPING STEP 4: Calculate Overlap Between Xenium Cells and ATAC Spots ===\n")

tryCatch({
  # Determine which spatial_info and feat_info to use based on available polygons
  available_polys <- getPolygonInfo(g_xenium)
  
  spatial_info_use <- if ("cell" %in% rownames(available_polys)) "cell" else NULL
  
  if (!is.null(spatial_info_use)) {
    g_xenium <- calculateOverlap(
      x = g_xenium,
      spatial_info = spatial_info_use,
      feat_info = "rna",
      name_overlap = "atac_overlap",
      verbose = TRUE
    )
    cat("✓ Overlap calculated using cell boundaries\n")
  } else {
    cat("⚠ No cell boundaries available - using centroid-based mapping\n")
    # Alternative: manually map cells to spots based on centroids
    xenium_meta <- pDataDT(g_xenium)
    
    # For each Xenium cell, find which ATAC spot contains it
    mapping_list <- lapply(seq_len(nrow(xenium_meta)), function(i) {
      cell_x <- xenium_meta$x[i]
      cell_y <- xenium_meta$y[i]
      
      # Check which ATAC polygon contains this point
      for (j in seq_len(nrow(atac_meta))) {
        if (abs(cell_x - atac_meta$x[j]) < dist && abs(cell_y - atac_meta$y[j]) < dist) {
          return(atac_meta$cell_ID[j])
        }
      }
      return(NA)
    })
    
    xenium_meta$atac_overlap <- unlist(mapping_list)
    cat("✓ Centroid-based mapping completed\n")
  }
  
}, error = function(e) {
  cat("⚠ Warning:", e$message, "\n")
})
```

```{r mapping-step5-extract-results, message=FALSE}
cat("\n=== MAPPING STEP 5: Extract and Analyze Mapping Results ===\n")

# Extract mapping results
mapping_results <- pDataDT(g_xenium)

# Find overlap/mapping column
overlap_col <- grep("atac|overlap", colnames(mapping_results), value = TRUE)[1]

if (is.na(overlap_col)) {
  cat("⚠ Warning: No overlap column found. Available columns:\n")
  print(colnames(mapping_results))
} else {
  cat("Using overlap column:", overlap_col, "\n")
  
  # Calculate density statistics
  density_stats <- mapping_results[
    !is.na(get(overlap_col)),
    .N,
    by = get(overlap_col)
  ]
  
  setnames(density_stats, c("atac_spot_id", "xenium_cells_per_spot"))
  setorder(density_stats, -xenium_cells_per_spot)
  
  # Calculate summary metrics
  total_mapped <- sum(!is.na(mapping_results[[overlap_col]]))
  total_unmapped <- sum(is.na(mapping_results[[overlap_col]]))
  occupied_spots <- nrow(density_stats)
  empty_spots <- nrow(atac_meta) - occupied_spots
  
  cat("\n=== MAPPING SUMMARY ===\n")
  cat("Total ATAC Spots:", nrow(atac_meta), "\n")
  cat("Total Xenium Cells:", nrow(mapping_results), "\n")
  cat("Xenium cells mapped to ATAC:", total_mapped, 
      sprintf("(%.1f%%)", 100 * total_mapped / nrow(mapping_results)), "\n")
  cat("Xenium cells unmapped:", total_unmapped, 
      sprintf("(%.1f%%)", 100 * total_unmapped / nrow(mapping_results)), "\n")
  cat("ATAC Spots with ≥1 Xenium cell:", occupied_spots, 
      sprintf("(%.1f%%)", 100 * occupied_spots / nrow(atac_meta)), "\n")
  cat("Empty ATAC Spots:", empty_spots, 
      sprintf("(%.1f%%)", 100 * empty_spots / nrow(atac_meta)), "\n")
  
  cat("\n=== CELL DENSITY DISTRIBUTION ===\n")
  print(summary(density_stats$xenium_cells_per_spot))
  
  cat("\nTop 10 densest ATAC spots:\n")
  print(head(density_stats, 10))
}
```

```{r mapping-step6-visualize, fig.width=14, fig.height=10, message=FALSE}
cat("\n=== MAPPING STEP 6: Visualize Mapping Results ===\n")

# Plot 1: Density histogram
p_density <- ggplot(density_stats, aes(x = xenium_cells_per_spot)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(xenium_cells_per_spot)), color = "red", linetype = "dashed", size = 1) +
  theme_minimal() +
  labs(
    title = "Xenium Cell Density Distribution per 10µm ATAC Spot",
    subtitle = sprintf("Mean: %.2f cells/spot | Median: %.2f cells/spot", 
                       mean(density_stats$xenium_cells_per_spot),
                       median(density_stats$xenium_cells_per_spot)),
    x = "Number of Xenium Cells per ATAC Spot",
    y = "Number of ATAC Spots",
    caption = "Expected: most spots have 0-2 cells given 10µm size and Xenium resolution"
  ) +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 10))

print(p_density)
ggsave(file.path(output.path, "04_mapping_density_histogram.png"), p_density, width = 10, height = 6, dpi = 150)
cat("✓ Saved: 04_mapping_density_histogram.png\n")

# Plot 2: Spatial distribution of mapped vs unmapped cells
if (atac_has_coords && xenium_has_coords) {
  mapping_results$mapping_status <- ifelse(is.na(mapping_results[[overlap_col]]), "Unmapped", "Mapped")
  
  p_spatial <- ggplot(mapping_results, aes(x = x, y = y, color = mapping_status)) +
    geom_point(size = 1.5, alpha = 0.6) +
    scale_color_manual(values = c("Mapped" = "green", "Unmapped" = "red")) +
    geom_point(data = atac_meta, aes(x = x, y = y), color = "blue", size = 2.5, 
               shape = 3, alpha = 0.4, stroke = 1, inherit.aes = FALSE) +
    coord_fixed() +
    theme_minimal() +
    labs(
      title = "Xenium Cells: Mapped vs Unmapped to ATAC Spots",
      subtitle = "Blue X = ATAC spot centroids",
      x = "X Coordinate (µm)",
      y = "Y Coordinate (µm)",
      color = "Status"
    ) +
    theme(plot.title = element_text(face = "bold"))
  
  print(p_spatial)
  ggsave(file.path(output.path, "05_mapping_spatial_distribution.png"), p_spatial, width = 10, height = 10, dpi = 150)
  cat("✓ Saved: 05_mapping_spatial_distribution.png\n")
}
```

```{r mapping-step7-save-results, message=FALSE}
cat("\n=== MAPPING STEP 7: Save Results ===\n")

# Save detailed mapping results
output_mapping <- file.path(output.path, "xenium_to_atac_mapping.csv")
fwrite(mapping_results, output_mapping)
cat("✓ Saved detailed mapping to:", output_mapping, "\n")

# Save density statistics
output_density <- file.path(output.path, "atac_spot_density_stats.csv")
fwrite(density_stats, output_density)
cat("✓ Saved density statistics to:", output_density, "\n")

# Create summary report
summary_report <- data.frame(
  Metric = c(
    "Total ATAC Spots",
    "Total Xenium Cells",
    "Xenium Cells Mapped",
    "Xenium Cells Unmapped",
    "ATAC Spots Occupied",
    "ATAC Spots Empty",
    "Mean Cells per Spot",
    "Median Cells per Spot",
    "Max Cells in Single Spot"
  ),
  Value = c(
    nrow(atac_meta),
    nrow(mapping_results),
    total_mapped,
    total_unmapped,
    occupied_spots,
    empty_spots,
    round(mean(density_stats$xenium_cells_per_spot), 2),
    round(median(density_stats$xenium_cells_per_spot), 2),
    max(density_stats$xenium_cells_per_spot)
  )
)

output_summary <- file.path(output.path, "mapping_summary_report.csv")
fwrite(as.data.table(summary_report), output_summary)
cat("✓ Saved summary report to:", output_summary, "\n")

cat("\n=== ALL OUTPUTS SAVED ===\n")
cat("Output directory:", output.path, "\n")
```

## Section 4: Post-Processing - Validation Checks







```{r validation-checks, message=FALSE}
cat("\n=== VALIDATION CHECKS ===\n")

# Check 1: Verify mapping coverage
if (exists("total_mapped") && exists("total_unmapped")) {
  mapping_pct <- 100 * total_mapped / nrow(mapping_results)
  if (mapping_pct > 90) {
    cat("✓ PASS: >90% of Xenium cells are mapped to ATAC spots\n")
  } else if (mapping_pct > 50) {
    cat("⚠ CAUTION: Only", round(mapping_pct, 1), "% of Xenium cells mapped\n")
  } else {
    cat("✗ FAIL: <50% of Xenium cells mapped (possible alignment issue)\n")
  }
}

# Check 2: Verify spot occupancy
if (exists("occupied_spots") && exists("empty_spots")) {
  occupancy_pct <- 100 * occupied_spots / nrow(atac_meta)
  if (occupancy_pct > 70) {
    cat("✓ PASS: >70% of ATAC spots have at least 1 Xenium cell\n")
  } else if (occupancy_pct > 30) {
    cat("⚠ CAUTION:", round(occupancy_pct, 1), "% of ATAC spots occupied\n")
  } else {
    cat("✗ FAIL: <30% of ATAC spots have cells (possible misalignment)\n")
  }
}

# Check 3: Verify cell density is reasonable
if (exists("density_stats") && nrow(density_stats) > 0) {
  mean_density <- mean(density_stats$xenium_cells_per_spot)
  if (mean_density > 0.5 && mean_density < 10) {
    cat("✓ PASS: Mean cell density", round(mean_density, 2), 
        "cells/spot is reasonable for 10µm spots\n")
  } else if (mean_density > 0) {
    cat("⚠ CAUTION: Mean density", round(mean_density, 2), 
        "cells/spot may indicate sub-optimal alignment\n")
  } else {
    cat("✗ FAIL: No cells mapped to any ATAC spot\n")
  }
}

cat("\n=== END OF ANALYSIS ===\n")
cat("Review diagnostic plots and CSV files in:", output.path, "\n")
cat("\nNext steps:\n")
cat("1. Review overlay plots (01_debug_overlay_all.png, 02_debug_atac_spots.png, 03_debug_xenium_cells.png)\n")
cat("2. Check mapping quality plots (04_mapping_density_histogram.png, 05_mapping_spatial_distribution.png)\n")
cat("3. Review CSV files for detailed mapping results and statistics\n")
cat("4. If alignment is poor, adjust MOSAICField parameters and re-run\n")
```

## Session Info

```{r session-info}
sessionInfo()
```
