---
title: "ArchR ATAC Seq analysis"
editor_options: 
  chunk_output_type: console
---


# Setting up
```{r}
.libPaths('/projectnb/paxlab/presh/ArchR_libs')
.libPaths()
```


```{r}
library(ArchR)
library(parallel)
set.seed(1)
addArchRLocking(locking = TRUE)
addArchRGenome("hg38")
```

# DEEP & LOWSEQ
Let's look at DeepSeq and Lowseq data 

<!-- ## Creating arrow files -->

<!-- ```{r} -->
<!-- addArchRThreads(threads = 16)  -->

<!-- inputFiles <- c("/projectnb/paxlab/DATA/DriesSpatial/Atlasxomics/D1942_deepseq/D01942_NG06549_ATv008_1/fragments.sort.bed.gz", "/projectnb/paxlab/DATA/DriesSpatial/Atlasxomics/D1942_lowseq/fragments.tsv.gz") -->
<!-- names(inputFiles) = c('deepseq', 'lowseq') -->

<!-- # dir.create('/projectnb/paxlab/presh/Data') -->
<!-- knitr::opts_knit$set(root.dir = "/projectnb/paxlab/presh/Data") -->


<!-- ``` -->

<!-- I did generate my own arrow file, but takes forever to run , using Yeting's arrow files that have been filtered for cells in the spatial coordinates for the project  -->

<!-- ArrowFiles <- createArrowFiles( -->
<!--   inputFiles = inputFiles, -->
<!--   sampleNames = names(inputFiles), -->
<!--   minTSS = 2, #Dont set this too high because you can always increase later -->
<!--   minFrags = 1e+04,  -->
<!--   maxFrags = 1e+06, -->
<!--   addTileMat = TRUE, -->
<!--   addGeneScoreMat = TRUE, -->
<!--   verbose = TRUE -->
<!-- ) -->

<!-- They also have doublet info added into it, you can use a filterratio to remove it later on! -->


## Creating an ArchR project 

```{r}
arrow.path = '/projectnb/paxlab/yeting/SpatialATACseq/results/preprocessed/D01942_filtered_ArchR_Output/ArrowFiles'
ArrowFiles = c(file.path(arrow.path, 'Deepseq.arrow'), file.path(arrow.path, 'Lowseq.arrow'))

# doubScores <- addDoubletScores(
#     input = ArrowFiles,
#     k = 10, #Refers to how many cells near a "pseudo-doublet" to count.
#     knnMethod = "UMAP", #Refers to the embedding to use for nearest neighbor search with doublet projection.
#     LSIMethod = 1
# )


deepseq.proj <- ArchRProject(
  ArrowFiles = ArrowFiles[1], 
  outputDirectory = "/projectnb/paxlab/presh/Data",
  copyArrows = F
)


lowseq.proj <- ArchRProject(
  ArrowFiles = ArrowFiles[2], 
  outputDirectory = "/projectnb/paxlab/presh/Data",
  copyArrows = F
)


combined.proj <- ArchRProject(
  ArrowFiles = ArrowFiles, 
  outputDirectory = "/projectnb/paxlab/presh/Data",
  copyArrows = F
)


```


## Plotting QC metrics 

```{r}
df <- getCellColData(deepseq.proj, select = c("log10(nFrags)", "TSSEnrichment"))

p <- ggPoint(
    x = df[,1], 
    y = df[,2], 
    colorDensity = TRUE,
    continuousSet = "sambaNight",
    xlabel = "Log10 Unique Fragments",
    ylabel = "TSS Enrichment",
    xlim = c(log10(500), quantile(df[,1], probs = 1)),
    ylim = c(0, quantile(df[,2], probs = 1))
) + geom_hline(yintercept = 4, lty = "dashed") + geom_vline(xintercept = 3, lty = "dashed")

p
```


```{r}
df <- getCellColData(lowseq.proj, select = c("log10(nFrags)", "TSSEnrichment"))

p <- ggPoint(
    x = df[,1], 
    y = df[,2], 
    colorDensity = TRUE,
    continuousSet = "sambaNight",
    xlabel = "Log10 Unique Fragments",
    ylabel = "TSS Enrichment",
    xlim = c(log10(500), quantile(df[,1], probs = 1)),
    ylim = c(0, quantile(df[,2], probs = 1))
) + geom_hline(yintercept = 4, lty = "dashed") + geom_vline(xintercept = 3, lty = "dashed")

p
```


```{r}
p1 <- plotGroups(
    ArchRProj = combined.proj, 
    groupBy = "Sample", 
    colorBy = "cellColData", 
    name = "TSSEnrichment",
  plotAs = "ridges",
    baseSize = 10
)
p1
```


```{r}
p3 <- plotGroups(
    ArchRProj = combined.proj, 
    groupBy = "Sample", 
    colorBy = "cellColData", 
  name = "log10(nFrags)",
  plotAs = "ridges",
    baseSize = 10
)

p3
```




## LSI 

```{r}
deepseq.proj2 <- addIterativeLSI(
    ArchRProj = deepseq.proj,
    useMatrix = "TileMatrix", 
    name = "IterativeLSI", 
    iterations = 2, 
    clusterParams = list( #See Seurat::FindClusters
        resolution = c(0.2), 
        sampleCells = 10000, 
        n.start = 10
    ), 
    varFeatures = 25000, 
    dimsToUse = 1:30
)


lowseq.proj2 <- addIterativeLSI(
    ArchRProj = lowseq.proj,
    useMatrix = "TileMatrix", 
    name = "IterativeLSI", 
    iterations = 2, 
    clusterParams = list( #See Seurat::FindClusters
        resolution = c(0.2), 
        sampleCells = 10000, 
        n.start = 10
    ), 
    varFeatures = 25000, 
    dimsToUse = 1:30
)
```

## Clustering 
```{r}
deepseq.proj2 <- addClusters(
    input = deepseq.proj2,
    reducedDims = "IterativeLSI",
    method = "Seurat",
    name = "Clusters",
    resolution = 0.8
)


lowseq.proj2 <- addClusters(
    input = lowseq.proj2,
    reducedDims = "IterativeLSI",
    method = "Seurat",
    name = "Clusters",
    resolution = 0.8
)
```


## UMAP 

```{r}
deepseq.proj2 <- addUMAP(
    ArchRProj = deepseq.proj2, 
    reducedDims = "IterativeLSI", 
    name = "UMAP", 
    nNeighbors = 30, 
    minDist = 0.5, 
    metric = "cosine"
)

lowseq.proj2 <- addUMAP(
    ArchRProj = lowseq.proj2, 
    reducedDims = "IterativeLSI", 
    name = "UMAP", 
    nNeighbors = 30, 
    minDist = 0.5, 
    metric = "cosine"
)
```


```{r}
p1 <- plotEmbedding(ArchRProj = deepseq.proj2, colorBy = "cellColData", name = "Clusters", embedding = "UMAP")
p1

p2 <-  plotEmbedding(ArchRProj = lowseq.proj2, colorBy = "cellColData", name = "Clusters", embedding = "UMAP")
p2

```


We should also probably mark cells from which tissue they come and add as metadata!


## Visualizing the tissue locations
```{r}
spatial_locs <- read.csv(file.path('/projectnb/paxlab/yeting/SpatialATACseq/data',
                                   "/tissue_positions_list.csv"),
                         header = FALSE)
colnames(spatial_locs) <- c("barcode", "in_tissue", "array_row", "array_col", "x_spatial", "y_spatial")

xy <- spatial_locs[, c("barcode", "x_spatial", "y_spatial")]
xy2 = spatial_locs[spatial_locs$in_tissue == 1, ]


plot(x = xy2$x_spatial, y = xy2$y_spatial, cex = 0.1) + abline(h = 4000)

# if anything is above y spatial = 4000 it is tissue 488 , else 487 

xy2$tissue = ifelse(xy2$y_spatial > 4000, 'Tiss_488B', 'Tiss_487')

table(xy2$tissue) 
# More cells in tissue 488B 
```
Tissue 488 has 11645 cells 

Also Xenium has data for 110809 cells, but also has a bigger tissue region.


Let's add this to the CellColData which tissue it comes from 

For Deepseq - 
```{r}
deepseq.meta = data.frame(deepseq.proj@cellColData)
deepseq.meta$barcode = gsub('-1','' ,gsub('Deepseq#', '',rownames(deepseq.meta)))
deepseq.meta2 = merge(deepseq.meta, xy2, by = 'barcode',all.x = T)
# order based on deepseq.meta
match_indices <- match(deepseq.meta$barcode, deepseq.meta2$barcode)
deepseq.meta2_ord = deepseq.meta2[match_indices, ]


# Put it back into the project 
deepseq.proj2 = addCellColData(deepseq.proj, data = deepseq.meta2_ord$tissue, name = 'tissue', cells = getCellNames(deepseq.proj))

# # Let's visualize UMAP 
# plotEmbedding(
#   ArchRProj = deepseq.proj3,
#   embedding = "UMAP",
#   colorBy = "cellColData",
#   name = "tissue",
#   size = 1,
#   sampleCells = NULL,
#   baseSize = 10,
#   plotAs = "points")
# 
# plotEmbedding(
#   ArchRProj = deepseq.proj3,
#   embedding = "UMAP",
#   colorBy = "cellColData",
#   name = "nFrags",
#   size = 1,
#   sampleCells = NULL,
#   baseSize = 10,
#   plotAs = "points")

```


For Lowseq - 
```{r}
lowseq.meta = data.frame(lowseq.proj@cellColData)
lowseq.meta$barcode = gsub('-1','' ,gsub('Lowseq#', '',rownames(lowseq.meta)))
lowseq.meta2 = merge(lowseq.meta, xy2, by = 'barcode',all.x = T)
# order based on deepseq.meta
match_indices <- match(lowseq.meta$barcode, lowseq.meta2$barcode)
lowseq.meta2_ord = lowseq.meta2[match_indices, ]


# Put it back into the project 
lowseq.proj2 = addCellColData(lowseq.proj, data = lowseq.meta2_ord$tissue, name = 'tissue', cells = getCellNames(lowseq.proj))

# # Let's visualize UMAP 
# plotEmbedding(
#   ArchRProj = lowseq.proj3,
#   embedding = "UMAP",
#   colorBy = "cellColData",
#   name = "tissue",
#   size = 1,
#   sampleCells = NULL,
#   baseSize = 10,
#   plotAs = "points")
# 
# plotEmbedding(
#   ArchRProj = lowseq.proj3,
#   embedding = "UMAP",
#   colorBy = "cellColData",
#   name = "nFrags",
#   size = 1,
#   sampleCells = NULL,
#   baseSize = 10,
#   plotAs = "points")


```


Make plots of nFrags keeping the spatial locations 

```{r}


ggplot(deepseq.meta2_ord, aes(x = x_spatial, y = y_spatial, color = nFrags)) +
  geom_point(size = 0.6) +
  scale_color_viridis_c() + theme_classic()

ggplot(deepseq.meta2_ord, aes(x = x_spatial, y = y_spatial, color = TSSEnrichment)) +
  geom_point(size = 0.6) +
  scale_color_viridis_c() + theme_classic()



ggplot(lowseq.meta2_ord, aes(x = x_spatial, y = y_spatial, color = nFrags)) +
  geom_point(size = 0.6) +
  scale_color_viridis_c() + theme_classic()

ggplot(lowseq.meta2_ord, aes(x = x_spatial, y = y_spatial, color = TSSEnrichment)) +
  geom_point(size = 0.6) +
  scale_color_viridis_c() + theme_classic()
```






## Gene score correlations 
Date: 15 Sept 2025 

Load in the deepseq proj 
```{r}
# deepseq.proj = readRDS('/projectnb/paxlab/presh/Data/Save-ArchR-Project.rds')


# subset the cells that we look at the spatial tissue positions csv from
deepseq.meta = data.frame(deepseq.proj@cellColData)
deepseq.meta$barcode = gsub('-1','' ,gsub('Deepseq#', '',rownames(deepseq.meta)))
deepseq.meta2 = merge(deepseq.meta, xy2, by = 'barcode',all.x = T)
# order based on deepseq.meta
match_indices <- match(deepseq.meta$barcode, deepseq.meta2$barcode)
deepseq.meta2_ord = deepseq.meta2[match_indices, ]

# Put it back into the project 
deepseq.proj2 = addCellColData(deepseq.proj, data = deepseq.meta2_ord$tissue, name = 'tissue', cells = getCellNames(deepseq.proj))



lowseq.meta = data.frame(lowseq.proj@cellColData)
lowseq.meta$barcode = gsub('-1','' ,gsub('Lowseq#', '',rownames(lowseq.meta)))
lowseq.meta2 = merge(lowseq.meta, xy2, by = 'barcode',all.x = T)
# order based on deepseq.meta
match_indices <- match(lowseq.meta$barcode, lowseq.meta2$barcode)
lowseq.meta2_ord = lowseq.meta2[match_indices, ]

# Put it back into the project 
lowseq.proj2 = addCellColData(lowseq.proj, data = lowseq.meta2_ord$tissue, name = 'tissue', cells = getCellNames(lowseq.proj))


###### Getting gene scores 
gene_score_deepseq <- getMatrixFromProject(
  ArchRProj = deepseq.proj2,
  useMatrix = "GeneScoreMatrix"
)

gene_score_lowseq <- getMatrixFromProject(
  ArchRProj = lowseq.proj2,
  useMatrix = "GeneScoreMatrix"
)

# let's see if the cells and the fragments match
mat1 = gene_score_deepseq@assays@data$GeneScoreMatrix
mat2 = gene_score_lowseq@assays@data$GeneScoreMatrix

sum(gene_score_lowseq@elementMetadata ==  gene_score_deepseq@elementMetadata) == nrow(gene_score_deepseq@elementMetadata)

gene_score_deepseq_barcode = gsub('-1','' ,gsub('Deepseq#', '',rownames(gene_score_deepseq@colData)))
gene_score_lowseq_barcode = gsub('-1','' ,gsub('Lowseq#', '',rownames(gene_score_lowseq@colData)))

common_cols = intersect(gene_score_deepseq_barcode, gene_score_lowseq_barcode)



mat1_sub <- mat1[, paste('Deepseq#',common_cols, '-1',sep=''), drop = FALSE]
mat2_sub <- mat2[, paste('Lowseq#',common_cols,'-1', sep=''), drop = FALSE]



```

Making input for paste3
```{r}
library(data.table)
paste3_input = '/projectnb/paxlab/presh/paste3_input'


rownames(xy2) = xy2$barcode
xy3 = xy2[common_cols, ]
xy.488 = xy3[xy3$tissue == 'Tiss_488B',]

# fwrite(xy.488[,c('x_spatial', "y_spatial")], file.path(paste3_input,"slice1_coor.csv"), col.names = F, row.names = F)

rownames(mat1_sub) = gene_score_lowseq@elementMetadata$name
mat1_sub.488 = mat1_sub[,  paste('Deepseq#',xy.488$barcode, '-1',sep='')]

mat1_sub.488.t = t(mat1_sub.488)

# fwrite(as.matrix(mat1_sub.488.t), file = file.path(paste3_input,"slice1.csv"), row.names = F)

```


## Make featureplots 

Get genes that we like and look at the gene expression 

```{r}
int.genes = c('EPCAM', 'KRT5', 'EGFR', 'CD3E', 'CD8A', 'FOXP3', 'CD68', 'CD19', 'ACTA2', 'PECAM1', 'TOP2A', 'BRCA1')

rownames(mat1_sub.488.t) = gsub('-1','' ,gsub('Deepseq#', '',rownames(mat1_sub.488.t)))

(mat1_sub.488.t[,int.genes])

deepseq.meta2_ord.genes = merge(deepseq.meta2_ord, as.matrix(mat1_sub.488.t[,int.genes]), by.x = 'barcode', by.y = 'row.names')


ggplot(deepseq.meta2_ord.genes, aes(x = x_spatial, y = y_spatial, color = BRCA1)) +
  geom_point(size = 0.8) + theme_classic() +
  scale_color_gradientn(
    colors = c("blue", "white", "red"), # Define the colors for the gradient
    limits = c(0, 1), # Explicitly set the legend's data range from 0 to 1
    breaks = seq(0, 1, by = 0.5))
```



# DEEPSEQ

```{r}
addArchRThreads(threads = 16) 

inputFiles <- c("/projectnb/paxlab/DATA/DriesSpatial/Atlasxomics/D1942_deepseq/D01942_NG06549_ATv008_1/fragments.sort.bed.gz")
names(inputFiles) = c('deepseq')

# dir.create('/projectnb/paxlab/presh/Data')
knitr::opts_knit$set(root.dir = "/projectnb/paxlab/presh/Data")

arrow.path = '/projectnb/paxlab/yeting/SpatialATACseq/results/preprocessed/D01942_filtered_ArchR_Output/ArrowFiles'
ArrowFiles = c(file.path(arrow.path, 'Deepseq.arrow'))


deepseq.proj <- ArchRProject(
  ArrowFiles = ArrowFiles, 
  outputDirectory = "/projectnb/paxlab/presh/Data",
  copyArrows = F
)

# gene_score_deepseq <- getMatrixFromProject(
  # ArchRProj = deepseq.proj,
  # useMatrix = "GeneScoreMatrix"
# )

# saveRDS(gene_score_deepseq, '/projectnb/paxlab/presh/gene_score_deepseq.rds')


gene_score_deepseq = readRDS('/projectnb/paxlab/presh/gene_score_deepseq.rds')
```


```{r}
spatial_locs <- read.csv(file.path('/projectnb/paxlab/yeting/SpatialATACseq/data',
                                   "/tissue_positions_list.csv"),
                         header = FALSE)
colnames(spatial_locs) <- c("barcode", "in_tissue", "array_row", "array_col", "x_spatial", "y_spatial")

xy <- spatial_locs[, c("barcode", "x_spatial", "y_spatial")]
xy2 = spatial_locs[spatial_locs$in_tissue == 1, ]


plot(x = xy2$x_spatial, y = xy2$y_spatial, cex = 0.1) + abline(h = 4000)

# if anything is above y spatial = 4000 it is tissue 488 , else 487 

xy2$tissue = ifelse(xy2$y_spatial > 4000, 'Tiss_488B', 'Tiss_487')

table(xy2$tissue) 
```

```{r}
# subset the cells that we look at the spatial tissue positions csv from
deepseq.meta = data.frame(deepseq.proj@cellColData)
deepseq.meta$barcode = gsub('-1','' ,gsub('Deepseq#', '',rownames(deepseq.meta)))
deepseq.meta2 = merge(deepseq.meta, xy2, by = 'barcode',all.x = T)
# order based on deepseq.meta
match_indices <- match(deepseq.meta$barcode, deepseq.meta2$barcode)
deepseq.meta2_ord = deepseq.meta2[match_indices, ]

# Put it back into the project 
deepseq.proj2 = addCellColData(deepseq.proj, data = deepseq.meta2_ord$tissue, name = 'tissue', cells = getCellNames(deepseq.proj))

```

```{r}
ggplot(deepseq.meta2_ord, aes(x = x_spatial, y = y_spatial, color = nFrags)) +
  geom_point(size = 0.6) +
  scale_color_viridis_c() + theme_classic()
```

Let's get rid of edge effects
```{r}
hist(deepseq.meta2_ord$nFrags, breaks = 100)

deepseq.meta2_ord[deepseq.meta2_ord$nFrags > 5e5,]
sum(deepseq.meta2_ord$nFrags > 4e5)

int.cells = deepseq.meta2_ord[deepseq.meta2_ord$nFrags < 2e5,'barcode']

ggplot(deepseq.meta2_ord[deepseq.meta2_ord$barcode %in% int.cells, ], aes(x = x_spatial, y = y_spatial, color = nFrags)) +
  geom_point(size = 0.6) +
  scale_color_viridis_c() + theme_classic()


int.barcodes = paste('Deepseq#',int.cells, '-1',sep='')

# deepseq.proj3 = subsetArchRProject(
#   ArchRProj = deepseq.proj2,
#   cells = int.barcodes, force = T)

# Let's also keep only 488B tissue 
deepseq.meta3 = deepseq.meta2_ord[deepseq.meta2_ord$barcode %in% int.cells, ]
cells.488B = deepseq.meta3[deepseq.meta3$tissue == 'Tiss_488B', 'barcode']
barcodes.488B = paste('Deepseq#',cells.488B, '-1',sep='')
rownames(deepseq.meta3) = barcodes.488B

ggplot(deepseq.meta2_ord[deepseq.meta2_ord$barcode %in% cells.488B, ], aes(x = x_spatial, y = y_spatial, color = nFrags)) +
  geom_point(size = 0.6) +
  scale_color_viridis_c() + theme_classic()

ggplot(deepseq.meta2_ord[deepseq.meta2_ord$barcode %in% cells.488B, ], aes(x = x_spatial, y = y_spatial, color = TSSEnrichment)) +
  geom_point(size = 0.6) +
  scale_color_viridis_c() + theme_classic()


# tiss_488B.proj = subsetArchRProject(
#   ArchRProj = deepseq.proj2, outputDirectory = 'Deepseq_488B',
#   cells = barcodes.488B, force = T, dropCells = T)


tiss_488B.proj = deepseq.proj2[barcodes.488B, ]

write.csv(cells.488B, 'int_cells_trim_atac.csv')



```

## LSI and UMAP 

```{r}
tiss_488B.proj <- addIterativeLSI(
    ArchRProj = tiss_488B.proj,
    useMatrix = "TileMatrix", 
    name = "IterativeLSI", 
    iterations = 2, 
    clusterParams = list( #See Seurat::FindClusters
        resolution = c(0.2), 
        sampleCells = 10000, 
        n.start = 10
    ), 
    varFeatures = 25000, 
    dimsToUse = 1:30
)

# also add Iterative LSI with genescore matrix
tiss_488B.proj <- addIterativeLSI(
    ArchRProj = tiss_488B.proj,
    useMatrix = "GeneScoreMatrix", 
    name = "GeneScoreLSI", 
    iterations = 2, 
    clusterParams = list( #See Seurat::FindClusters
        resolution = c(0.2), 
        sampleCells = 10000, 
        n.start = 10
    ), 
    varFeatures = 5000, 
    dimsToUse = 1:30
)



tiss_488B.proj <- addClusters(
    input = tiss_488B.proj,
    reducedDims = "IterativeLSI",
    method = "Seurat",
    name = "Clusters",
    resolution = 0.5,
    force = TRUE
)

tiss_488B.proj <- addClusters(
    input = tiss_488B.proj,
    reducedDims = "GeneScoreLSI",
    method = "Seurat",
    name = "Clusters2",
    resolution = 0.5,
    force = TRUE
)


tiss_488B.proj <- addUMAP(
    ArchRProj = tiss_488B.proj, 
    reducedDims = "IterativeLSI", 
    name = "UMAP", 
    nNeighbors = 30, 
    minDist = 0.5, 
    metric = "cosine", 
    force = T
)

tiss_488B.proj <- addUMAP(
    ArchRProj = tiss_488B.proj, 
    reducedDims = "GeneScoreLSI", 
    name = "UMAP2", 
    nNeighbors = 30, 
    minDist = 0.5, 
    metric = "cosine", 
    force = T
)

plotEmbedding(ArchRProj = tiss_488B.proj, colorBy = "cellColData", name = "Clusters", embedding = "UMAP")
plotEmbedding(ArchRProj = tiss_488B.proj, colorBy = "cellColData", name = "nFrags", embedding = "UMAP")
plotEmbedding(ArchRProj = tiss_488B.proj, colorBy = "cellColData", name = "TSSEnrichment", embedding = "UMAP")



plotEmbedding(ArchRProj = tiss_488B.proj, colorBy = "cellColData", name = "Clusters", embedding = "UMAP2")
plotEmbedding(ArchRProj = tiss_488B.proj, colorBy = "cellColData", name = "nFrags", embedding = "UMAP2")
plotEmbedding(ArchRProj = tiss_488B.proj, colorBy = "cellColData", name = "TSSEnrichment", embedding = "UMAP2")

```


```{r}
cluster.meta = as.data.frame(tiss_488B.proj@cellColData)
cluster.meta$barcode = gsub('-1','' ,gsub('Deepseq#', '',rownames(cluster.meta)))

deepseq.meta3 = merge(deepseq.meta2_ord[,c('barcode', 'x_spatial', 'y_spatial')], cluster.meta, by = 'barcode')

ggplot(deepseq.meta3, aes(x = x_spatial, y = y_spatial, color = Clusters)) +
  geom_point(size = 0.9) +
 theme_classic() + scale_color_manual(values = c('C1' = 'darkgreen', 'C2' = 'blue', 'C4' = 'pink', 'C3' = 'yellow'))

```

```{r}
# update the meta data in the ArchR project 
rownames(deepseq.meta3) =  paste('Deepseq#',deepseq.meta3$barcode, '-1',sep='')
# order by archr proj cell names 
deepseq.meta3_ord = deepseq.meta3[rownames(getCellColData(tiss_488B.proj)), ]

tiss_488B.proj2 = addCellColData(tiss_488B.proj, deepseq.meta3_ord$x_spatial, name = 'x_spatial', cells = getCellNames(tiss_488B.proj))
tiss_488B.proj2 = addCellColData(tiss_488B.proj2, deepseq.meta3_ord$y_spatial, name = 'y_spatial', cells = getCellNames(tiss_488B.proj2))


saveArchRProject(tiss_488B.proj2, '/projectnb/paxlab/presh/Data')

```




## Finding markers for each cluster 

```{r}
markersGS <- getMarkerFeatures(
    ArchRProj = tiss_488B.proj, 
    useMatrix = "GeneScoreMatrix", 
    groupBy = "Clusters",
    bias = c("TSSEnrichment", "log10(nFrags)"),
    testMethod = "wilcoxon"
)


markerList <- getMarkers(markersGS, cutOff = "FDR <= 0.1 & Log2FC >= 0.5")
markerList
```

```{r}
tiss_488B.proj <- addImputeWeights(tiss_488B.proj)


markerGenes  <- c(
    "CD34",  #Early Progenitor
    "GATA1", #Erythroid
    "PAX5", "MS4A1", "MME", #B-Cell Trajectory
    "CD14", "MPO", #Monocytes
    "CD3D", "CD8A"#TCells
  )

p = plotEmbedding(
    ArchRProj = tiss_488B.proj, 
    colorBy = "GeneScoreMatrix", 
    name = markerGenes, 
    embedding = "UMAP",
    imputeWeights = getImputeWeights(tiss_488B.proj)
)



p2 <- lapply(p, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 3),p2))
```






## Epianeufinder results 


### Reading in hg38 metadata 
```{r}
library(data.table)
library(dplyr)
library(tidyr)

hg38.chrom <- fread("http://hgdownload.cse.ucsc.edu/goldenpath/hg38/database/cytoBand.txt.gz", 
           col.names = c("chrom","chromStart","chromEnd","name","gieStain"))

hg38.chrom.v2 = hg38.chrom[hg38.chrom$chrom %in% paste0('chr', 1:22), ]


hg38.chrom.v3 <- hg38.chrom.v2 %>%
  mutate(
    arm = ifelse(startsWith(name, "p"), "p", "q"),
    chrom_arm = paste0(chrom, arm)
  )


arm_ranges <- hg38.chrom.v3 %>%
  group_by(chrom, arm = ifelse(startsWith(name, "p"), "p", "q")) %>%
  summarize(
    arm_start = min(chromStart),
    arm_end = max(chromEnd),
    .groups = "drop"
  ) %>%
  mutate(chrom_arm = paste0(chrom, arm))

```


### Annotating arm bands to epianeufinder results 

From the paper:
In summary, epiAneufinder takes scATAC-seq BAM files or 10x fragment files as input, and outputs a RDS file and a TSV file with the identified copy number states for each bin per cell, labeled as “loss” (0), “normal” (1) or “gain” (2). 

```{r}


cnv_calls <- fread('/projectnb/paxlab/presh/Data/epianeufinder/atac_results/epiAneufinder_results/results_table.tsv')


cnv_calls$start <- as.numeric(cnv_calls$start)
cnv_calls$end   <- as.numeric(cnv_calls$end)
arm_ranges$arm_start <- as.numeric(arm_ranges$arm_start)
arm_ranges$arm_end   <- as.numeric(arm_ranges$arm_end)

calls_annotated <- cnv_calls %>%
  rowwise() %>%
  mutate(
    arm = arm_ranges$arm[
      seq == arm_ranges$chrom &
      start >= arm_ranges$arm_start &
      end <= arm_ranges$arm_end
    ],
    chrom_arm = arm_ranges$chrom_arm[
      seq == arm_ranges$chrom &
      start >= arm_ranges$arm_start &
      end <= arm_ranges$arm_end
    ]
  ) %>%
  ungroup()

# rearrange the cols 
calls_annotated <- calls_annotated %>%
  select(V1, seq, start, end, arm, chrom_arm, everything())
```

### Summarizing the cnv calls 
```{r}
calls_long <- calls_annotated %>%
  pivot_longer(
    cols = matches("^cell-"),        # Adjust if your columns are named differently
    names_to = "cell",
    values_to = "value"
  )

chrom_arm_summary <- calls_long %>%
  group_by(chrom_arm, cell, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(chrom_arm, cell) %>%
  mutate(prop = count / sum(count)) %>%
  ungroup()


summary_text <- chrom_arm_summary %>%
  arrange(chrom_arm, cell, value) %>%
  group_by(chrom_arm, cell) %>%
  summarise(summary = paste0(
    round(100*prop, 1), "% ", value, collapse = ", "
  ), .groups = "drop")



summary_wide <- chrom_arm_summary %>%
  select(chrom_arm, cell, value, prop) %>%
  pivot_wider(
    names_from = value,
    values_from = prop,
    values_fill = 0
  )

### Making the data matrix wide 
# 1. Gather into long format by 'value'
df_long <- summary_wide %>%
  pivot_longer(cols = c("0", "1", "2"), names_to = "value", values_to = "prop")

# 2. Make wide matrix
df_wide <- df_long %>%
  mutate(arm_value = paste0(chrom_arm, "_", value)) %>%
  select(cell, arm_value, prop) %>%
  pivot_wider(names_from = arm_value, values_from = prop, values_fill = 0)

# 3. Result: cells as rows, columns like chr10p_0, chr10p_1, chr10p_2, chr10q_0, etc.
mat <- as.matrix(df_wide[,-1])    # To get just the matrix (without cell names)
rownames(mat) <- df_wide$cell    # Set rownames as cell

# If you want the data frame with cell as first column:
final_df <- df_wide

```


### Add spatial coordinates to this - 

Load in ArchR project 
```{r}
tiss_488B.proj = loadArchRProject(path = "/projectnb/paxlab/presh/Data", force = FALSE, showLogo = TRUE)
```


```{r}
cluster.meta = as.data.frame(tiss_488B.proj@cellColData)
cluster.meta$barcode = gsub('Deepseq#', 'cell-',rownames(cluster.meta))

deepseq.meta3 = merge(final_df, cluster.meta, by.y = 'barcode', by.x = 'cell')

ggplot(deepseq.meta3, aes(x = x_spatial, y = y_spatial, color = chr1q_2)) +
  geom_point(size = 0.9) +
 theme_classic() + scale_color_gradient(low = "grey", high = "red")


ggplot(deepseq.meta3, aes(x = x_spatial, y = y_spatial, color = chr8q_2)) +
  geom_point(size = 0.9) +
 theme_classic() + scale_color_gradient(low = "grey", high = "red")


ggplot(deepseq.meta3, aes(x = x_spatial, y = y_spatial, color = chr7q_0)) +
  geom_point(size = 0.9) +
 theme_classic() + scale_color_gradient(low = "grey", high = "blue")

ggplot(deepseq.meta3, aes(x = x_spatial, y = y_spatial, color = chr21q_0)) +
  geom_point(size = 0.9) +
 theme_classic() + scale_color_gradient(low = "grey", high = "blue")
```












