---
title: "CNV_calling_ATAC"
editor_options: 
  chunk_output_type: console
---

Did CNV calling using these 2 tools

# Epianeufinder
## Running the tool

Opened a VS code with 2 cores! Gets killed with 1 core
Also in case the session gets terminated, no need to delete and restart. epianeufinder resumes from where it stopped 

###For deepseq data - 
```{bash}
library(epiAneufinder)

cd /projectnb/paxlab/presh/Data
awk 'NR==FNR{a[$1]; next} $4 in a' barcodes.txt fragments.sort.bed > fragments.sort.filtered.bed

epiAneufinder(input="/projectnb/paxlab/presh/Data/fragments.sort.filtered.bed", #Enter path to your fragments.tsv file or the folder containing bam files
              outdir="/projectnb/paxlab/presh/Data/epianeufinder/atac_results", #Path to the directory where results should be written 
              blacklist="/projectnb/paxlab/presh/Data/epianeufinder/hg38-blacklist.v2.bed", #Path to bed file that contains the blacklisted regions of your genome
              windowSize=1e6, 
              genome="BSgenome.Hsapiens.UCSC.hg38", #Substitute with relevant BSgenome
              exclude=c('chrX','chrY','chrM'), 
              reuse.existing=TRUE,
              title_karyo="Karyogram of sample data", 
              ncores=4,
              minFrags=20000,
              minsizeCNV=0,
              k=4,
              plotKaryo=TRUE)
```


###For lowseq data - 

Opened a VS code with 2 cores! Gets killed with 1 core

Need to filter out the gz file /projectnb/paxlab/DATA/DriesSpatial/Atlasxomics/D1942_lowseq/fragments.tsv.gz 

zcat /projectnb/paxlab/DATA/DriesSpatial/Atlasxomics/D1942_lowseq/fragments.tsv.gz  | awk 'NR==FNR{a[$1]; next} $4 in a' /projectnb/paxlab/presh/Data/barcodes.txt - > 	/projectnb/paxlab/presh/Data/lowseq.fragments.sort.filtered.bed


cd /projectnb/paxlab/presh/Data
module load R 
R

```{r}
library(epiAneufinder)

epiAneufinder(input="/projectnb/paxlab/presh/Data/lowseq.fragments.sort.filtered.bed", #Enter path to your fragments.tsv file or the folder containing bam files
              outdir="/projectnb/paxlab/presh/Data/epianeufinder/lowseq_atac_results", #Path to the directory where results should be written 
              blacklist="/projectnb/paxlab/presh/Data/epianeufinder/hg38-blacklist.v2.bed", #Path to bed file that contains the blacklisted regions of your genome
              windowSize=1e6, 
              genome="BSgenome.Hsapiens.UCSC.hg38", #Substitute with relevant BSgenome
              exclude=c('chrX','chrY','chrM'), 
              reuse.existing=TRUE,
              title_karyo="Karyogram of lowseq data", 
              ncores=4,
              minFrags=20000,
              minsizeCNV=0,
              k=4,
              plotKaryo=TRUE)
```






## Reading in results 

### For deepseq - 
```{r}
options(scipen = 999)
library(data.table)
cnv_calls <- fread('/projectnb/paxlab/presh/Data/epianeufinder/atac_results/epiAneufinder_results/results_table.tsv')


cnv_calls$start <- as.numeric(cnv_calls$start)
cnv_calls$end   <- as.numeric(cnv_calls$end)
# arm_ranges$arm_start <- as.numeric(arm_ranges$arm_start)
# arm_ranges$arm_end   <- as.numeric(arm_ranges$arm_end)

rownames(cnv_calls) = paste(cnv_calls$seq, cnv_calls$start, cnv_calls$end, sep = '_')
cnv_calls2 = cnv_calls[,-c(1:4)]
rownames(cnv_calls2) = rownames(cnv_calls)

```


```{r}
library("distances")
# Compute distances between cells
dist_mat <- as.dist(distances(t(cnv_calls2)))

# Hierarchical clustering
library(fastcluster)
hc <- fastcluster::hclust(dist_mat, method = "ward.D")

# Plot dendrogram
# plot(hc, main="Hierarchical Clustering of Cells", xlab="Cells", sub="", cex=0.6)
```

```{r}
clusters <- cutree(hc, k = 6) 

cluster_counts <- table(clusters)
print(cluster_counts)

# Basic dendrogram with colored branches by cluster
library(dendextend)
dend <- as.dendrogram(hc)
dend_colored <- color_branches(dend, k = 6)

# plot(dend_colored, main = "Hierarchical Clustering Dendrogram (12 clusters)", ylab = "Height")

# Optionally, visualize cluster assignments on a heatmap
library(pheatmap)

# Assume cell_mat_t is your cell-by-bin matrix from above
annotation <- data.frame(Cluster = as.factor(clusters))
rownames(annotation) <- rownames(t(cnv_calls2))

# Suppose you have clusters numbered 1 to 10
cluster_levels <- sort(unique(annotation$Cluster))
cluster_colors <- setNames(RColorBrewer::brewer.pal(length(cluster_levels), "Set3"), cluster_levels)
annotation_colors <- list(Cluster = cluster_colors)
```


####Plotting the clusters in the heatmap 
```{r}
my_heatmap <-pheatmap(t(cnv_calls2), annotation_row = annotation,
         show_rownames = FALSE, cluster_rows = hc, cluster_cols = FALSE, cutree_rows = 6,
         main = "Cell-by-Bin Heatmap, colored by 6 clusters",
         annotation_colors = annotation_colors)


save_pheatmap <- function(x, filename, width=16, height=8){
  stopifnot(!missing(x))
  stopifnot(!missing(filename))
  if(grepl(".png",filename)){
    png(filename, width=width, height=height, units = "in", res=300)
    grid::grid.newpage()
    grid::grid.draw(x$gtable)
    dev.off()
  }
  else if(grepl(".pdf",filename)){
    pdf(filename, width=width, height=height)
    grid::grid.newpage()
    grid::grid.draw(x$gtable)
    dev.off()
  }
  else{
    print("Filename did not contain '.png' or '.pdf'")
  }
}


save_pheatmap(my_heatmap, "/projectnb/paxlab/presh/Data/epianeufinder/cnv_6clust_pheatmap.png")
dev.off()
```

####To determine the number of clusters 
```{r}
last_heights <- tail(hc$height, 30)
num_clusters <- 31:2 #how many clusters do we want to evaluate?

#Plot the last 30 values against the range 2-31
plot(num_clusters, last_heights, type = 'o', col = 'blue', pch = 16,
     main = 'Cluster Distance vs Number of Clusters',
     xlab = 'Number of Clusters', ylab = 'Cluster Distance')

#Add a red dot at the elbow point
points(num_clusters[28], last_heights[28], col = 'red', pch = 16)
```


####Plot clusters spatially 
```{r}
.libPaths('/projectnb/paxlab/presh/ArchR_libs')
.libPaths()

library(ArchR)

# tiss_488B.proj = loadArchRProject(path = "/projectnb/paxlab/presh/Data", force = FALSE, showLogo = TRUE)

cluster.meta = as.data.frame(tiss_488B.proj2@cellColData)
cluster.meta$barcode = gsub('Deepseq#', 'cell-',rownames(cluster.meta))

# add cluster coordinates from hclust 
cluster.meta2 = merge(cluster.meta, annotation, by.x = 'barcode', by.y = 'row.names')
# cluster.meta2$Cluster = as.character(cluster.meta2$Cluster)

.libPaths("/usr3/graduate/preshita/R/x86_64-pc-linux-gnu-library/4.4")
library(ggplot2)

annotation_colors_cluster_palette <- c(
  "1" = "#8DD3C7", "2" = "#FFFFB3", "3" = "#BEBADA", "4" = "#FB8072", 
  "5" = "#80B1D3", "6" = "#FDB462"
)

ggplot(cluster.meta2, aes(x = x_spatial, y = y_spatial, color = Cluster)) +
  geom_point(size = 0.9) +
  theme_classic() +
  scale_color_manual(values = annotation_colors_cluster_palette) 

```


#### Making a barplot of the clusters between ArchR and hclust 

Subclone by ATAC cluster
```{r}

annotation_colors_cluster_palette <- c(
  "1" = "#8DD3C7", "2" = "#FFFFB3", "3" = "#BEBADA", "4" = "#FB8072", 
  "5" = "#80B1D3", "6" = "#FDB462"
)

# Ensure Cluster is a factor/character that matches palette keys
cluster.meta2$Cluster <- factor(cluster.meta2$Cluster, levels = names(annotation_colors_cluster_palette))

# Create the stacked bar plot
ggplot(cluster.meta2, aes(x = Clusters_tile, fill = Cluster)) +
  geom_bar(position = "stack", color = "black") +
  scale_fill_manual(values = annotation_colors_cluster_palette) +
  labs(x = "Clusters_tile", y = "Number of Cells", fill = "Cluster") +
  theme_classic()
```




### For lowseq - 
```{r}
options(scipen = 999)
library(data.table)
low_cnv_calls <- fread('/projectnb/paxlab/presh/Data/epianeufinder/lowseq_atac_results/epiAneufinder_results/results_table.tsv')


low_cnv_calls$start <- as.numeric(low_cnv_calls$start)
low_cnv_calls$end   <- as.numeric(low_cnv_calls$end)
# arm_ranges$arm_start <- as.numeric(arm_ranges$arm_start)
# arm_ranges$arm_end   <- as.numeric(arm_ranges$arm_end)

rownames(low_cnv_calls) = paste(low_cnv_calls$seq, low_cnv_calls$start, low_cnv_calls$end, sep = '_')
low_cnv_calls2 = low_cnv_calls[,-c(1:4)]
rownames(low_cnv_calls2) = rownames(low_cnv_calls)

```


### For deepseq - 
```{r}
options(scipen = 999)
library(data.table)
deep_cnv_calls <- fread('/projectnb/paxlab/presh/Data/epianeufinder/atac_results/epiAneufinder_results/results_table.tsv')


deep_cnv_calls$start <- as.numeric(deep_cnv_calls$start)
deep_cnv_calls$end   <- as.numeric(deep_cnv_calls$end)
# arm_ranges$arm_start <- as.numeric(arm_ranges$arm_start)
# arm_ranges$arm_end   <- as.numeric(arm_ranges$arm_end)

rownames(deep_cnv_calls) = paste(deep_cnv_calls$seq, deep_cnv_calls$start, deep_cnv_calls$end, sep = '_')
deep_cnv_calls2 = deep_cnv_calls[,-c(1:4)]
rownames(deep_cnv_calls2) = rownames(deep_cnv_calls)

```


### Comparing low and deepseq - 

so the windows are the same, cells are not, take common cells 
```{r}
sum(rownames(deep_cnv_calls2) == rownames(low_cnv_calls2)) == nrow(low_cnv_calls2)

common.cells = intersect(colnames(deep_cnv_calls2) , colnames(low_cnv_calls2)) 

deep_cnv_calls3 = deep_cnv_calls2[, ..common.cells]
low_cnv_calls3 = low_cnv_calls2[, ..common.cells]

M1_rect = deep_cnv_calls3
M2_rect = low_cnv_calls3

```



```{r}
cnv_method1_discrete <- as.matrix(M1_rect)
cnv_method2_discrete <- as.matrix(M2_rect)

num_genes = nrow(M1_rect)
num_cells = length(common.cells)

cat("--- Discrete Calls Comparison: Percent Agreement ---\n")
percent_agreement <- sum(cnv_method1_discrete == cnv_method2_discrete) / (num_genes * num_cells)
cat("Overall Percent Agreement:", round(percent_agreement * 100, 2), "%\n")

# You can also break this down by specific state agreement
# E.g., agreement on 'gain' calls (+1)
gain_agree <- sum(cnv_method1_discrete == 1 & cnv_method2_discrete == 1) / sum(cnv_method1_discrete == 1 | cnv_method2_discrete == 1)
cat("Jaccard Index for Gain Calls:", round(gain_agree, 4), "\n\n")
```

```{r}
# install.packages("caret") # For confusionMatrix
library(caret)

cat("--- Discrete Calls Comparison: Confusion Matrix (Method1 vs Method2) ---\n")
# Convert to factors
factor_m1 <- factor(as.vector(cnv_method1_discrete), levels = c(0,1,2))
factor_m2 <- factor(as.vector(cnv_method2_discrete), levels = c(0,1,2))

conf_matrix <- confusionMatrix(factor_m1, factor_m2, mode = "everything")
print(conf_matrix)
cat("\n")
```





# ATACNV
```{r}
library(AtaCNV)
```

```{r}
fragment_file <- "/projectnb/paxlab/presh/Data/fragments.sort.filtered.bed.gz"
cell_barcodes <- read.csv("/projectnb/paxlab/presh/Data/int_cells_trim_atac.csv", header=FALSE)

```


I don't want to generate the counts file by myself, using the epianeufinder results 

```{r}
counts_gc <- readRDS('/projectnb/paxlab/presh/Data/epianeufinder/atac_results/epiAneufinder_results/counts_gc_corrected.rds')
counts_summary <- readRDS('/projectnb/paxlab/presh/Data/epianeufinder/atac_results/epiAneufinder_results/count_summary.rds') # for info on ranges
```

```{r}
count = assay(counts_summary)
rownames(count) = paste(counts_summary@rowRanges$wSeq, counts_summary@rowRanges$wStart, counts_summary@rowRanges$wEnd, sep = '_')

norm_re <- normalize_AtaCNV(t(count),
                             genome = "hg38", 
                             mode = "all cells",
                             output_dir = "/projectnb/paxlab/presh/AtaCNV_output",
                             output_name = "norm_re.rds",
                             gc_correction = TRUE
                             )
```



```{r}
options(scipen = 999)
normalize_AtaCNV=  function (count, genome = "hg38", mode = "normal cells", thre_bin = 0.8, 
    thre_cell = 1, thre_mappability = 5e+05, logFT = F, dlm_dV = 0.3, 
    dlm_dW = 0.01, count_paired = "none", normal_cells = "none", 
    cell_cluster = "none", cluster_method = "kmeans", K = 5, 
    thre_correlation = 0.95, use_max_cor_cluster = F, thre_ncell = 30, 
    cutoff = 3, gc_correction = TRUE, output_dir = "./", output_name = "none") 
{
    count <- t(count)
    if (genome == "hg19") {
        data("bin_info_hg19")
        bin <- bin_info_hg19
    }
    else if (genome == "hg38") {
        data("bin_info_hg38")
        bin <- bin_info_hg38
    }
    else if (genome == "mm10") {
        data("bin_info_mm10")
        bin <- bin_info_mm10
    }
    use_paired <- F
    if (mode == "matched normal sample") {
        if (identical(count_paired, "none")) {
            print("Count matrix of paired normal sample should be provided")
        }
        use_paired <- T
        count_paired <- t(count_paired)
    }
    else if (mode == "normal cells") {
        if (identical(normal_cells, "none")) {
            print("Normal cells in this sample should be provided")
        }
    }
    
    # subset the bin to count bins 
    rownames(bin) <- paste(bin$chr, bin$start + 1, bin$end, sep = '_')
    bin <- bin[colnames(count),]
    
    print("Step 1: filtering bins and cells...")
    norm_count <- count
    if (use_paired) {
        norm_count <- cbind(count_paired, norm_count)
        thre_zero_bin <- thre_bin
        temp <- rowSums(norm_count <= 0)
        f1 <- temp < thre_zero_bin * dim(norm_count)[2] & bin$map >= 
            thre_mappability
        print(paste(" number of bin before filter:", length(f1)))
        print(paste(" number of bin after filter:", sum(f1)))
        norm_count <- norm_count[f1, ]
        q <- quantile(norm_count, probs = c(1:100) * 0.01)
        thre <- q[99]
        norm_count[norm_count > thre] <- thre
        thre_zero_cell <- thre_cell
        temp <- colSums(norm_count[, (ncol(count_paired) + 1):ncol(norm_count)] <= 
            0)
        f2 <- temp < thre_zero_cell * nrow(norm_count)
        print(paste(" number of cells before filter:", length(f2)))
        print(paste(" number of cells after filter:", sum(f2)))
        norm_count <- cbind(count_paired[f1, ], count[f1, f2])
    }
    else {
        thre_zero_bin <- thre_bin
        temp <- rowSums(norm_count <= 0)
        f1 <- temp < thre_zero_bin * dim(norm_count)[2] & bin$map >= 
            thre_mappability
        print(paste(" number of bin before filter:", length(f1)))
        print(paste(" number of bin after filter:", sum(f1)))
        norm_count <- norm_count[f1, ]
        q <- quantile(norm_count, probs = c(1:100) * 0.01)
        thre <- q[99]
        norm_count[norm_count > thre] <- thre
        thre_zero_cell <- thre_cell
        temp <- colSums(norm_count <= 0)
        f2 <- temp < thre_zero_cell * nrow(norm_count)
        print(paste(" number of cells before filter:", length(f2)))
        print(paste(" number of cells after filter:", sum(f2)))
        norm_count <- count[f1, f2]
    }
    if (!identical(normal_cells, "none")) {
        normal_cells <- intersect(colnames(count)[f2], colnames(count[, 
            normal_cells]))
    }
    if (!identical(cell_cluster, "none")) {
        cell_cluster <- cell_cluster[f2]
    }
    print("Step 2: smoothing read count signals...")
    if (logFT) {
        norm_count <- log(sqrt(norm_count) + sqrt(norm_count + 
            1))
    }
    else {
        norm_count <- log(norm_count + 1)
    }
    dlm_sm <- function(y) {
        model <- dlm::dlmModPoly(order = 1, dV = dlm_dV, dW = dlm_dW)
        x <- dlm::dlmSmooth(y, model)$s
        x <- x[2:length(x)]
        return(x)
    }
    dlm_sm_by_chr <- function(y) {
        bin_ <- bin[f1, ]
        model <- dlm::dlmModPoly(order = 1, dV = dlm_dV, dW = dlm_dW)
        x <- 1:length(y)
        for (i in unique(bin_$chr)) {
            f_ <- bin_$chr == i
            if (sum(f_) > 2) {
                x_ <- dlm::dlmSmooth(y[f_], model)$s
                x[f_] <- x_[-1]
            }
            else {
                x[f_] <- y[f_]
            }
        }
        return(x)
    }
    temp <- apply(norm_count, 2, dlm_sm)
    norm_count <- exp(temp)
    norm_count <- edgeR::cpm(norm_count)
    norm_count <- log(norm_count + 1)
    q <- quantile(norm_count, probs = c(1:100) * 0.01)
    thre <- q[99]
    norm_count[norm_count > thre] <- thre
    if (use_paired) {
        norm_count_paired <- norm_count[, 1:ncol(count_paired)]
        norm_count <- norm_count[, (ncol(count_paired) + 1):ncol(norm_count)]
    }
    else {
        norm_count <- norm_count
    }
    print("Step 3: calculating baseline of normal cells...")
    baseline <- data.frame(n = c(1:nrow(norm_count)))
    baseline$med_all <- apply(norm_count, 1, median)
    if (mode == "normal cells") {
        baseline$med_paired <- apply(norm_count[, normal_cells], 
            1, median)
        baseline$selected_normal <- baseline$med_paired
        print(" Baseline is calculated using median read count of normal cells")
    }
    if (mode == "all cells") {
        baseline$med_paired <- baseline$med_all
        baseline$selected_normal <- baseline$med_all
        print(" Baseline is calculated using median read count of all cells")
    }
    if (mode == "none" | mode == "matched normal sample") {
        if (identical(cell_cluster, "none")) {
            print(" Cell cluster not provided. Assigning cell clusters...")
            if (cluster_method == "hc") {
                d <- parallelDist::parallelDist(t(norm_count))
                hc_re <- hclust(d, method = "ward.D")
                cluster <- cutree(hc_re, k = K)
            }
            if (cluster_method == "kmeans") {
                temp <- irlba::prcomp_irlba(t(norm_count), n = 50)$x
                cluster <- kmeans(temp, centers = K)$cluster
            }
        }
        else {
            cluster <- as.factor(cell_cluster)
            K <- length(unique(cluster))
        }
    }
    if (mode == "none") {
        print(" Selecting normal cells using correlation with gene counts...")
        baseline$med_paired <- bin$gene[f1]
        corr <- c(1:K)
        for (i in 1:K) {
            j <- unique(cluster)[i]
            if (sum(cluster == j) == 1) {
                baseline[, i + 3] <- norm_count[, cluster == 
                  j]
            }
            else {
                baseline[, i + 3] <- apply(norm_count[, cluster == 
                  j], 1, median)
            }
            corr_ <- cor(baseline[, i + 3], baseline$med_paired, 
                method = "pearson")
            corr[i] <- corr_
            if (sum(cluster == j) < thre_ncell) {
                corr[i] <- -10
            }
            print(paste0(" cluster ", j, ";", sum(cluster == 
                j), "cells;", "corr=", corr_))
        }
        i <- which.max(corr)
        j <- unique(cluster)[i]
        baseline$selected_normal <- apply(norm_count[, cluster == 
            j], 1, median)
        print(paste0(" selected normal: ", sum(cluster == j), 
            " cells"))
        print(paste0(" max correlation to gene counts: ", corr[i]))
        print(" Baseline is calculated using median read count of selected normal cells")
    }
    if (mode == "matched normal sample") {
        print(" Selecting normal cells using paired normal sample...")
        baseline$med_paired <- apply(norm_count_paired, 1, median)
        corr <- c(1:K)
        for (i in 1:K) {
            j <- unique(cluster)[i]
            if (sum(cluster == j) == 1) {
                baseline[, i + 3] <- norm_count[, cluster == 
                  j]
            }
            else {
                baseline[, i + 3] <- apply(norm_count[, cluster == 
                  j], 1, median)
            }
            corr_ <- cor(baseline[, i + 3], baseline$med_paired, 
                method = "pearson")
            corr[i] <- corr_
            if (sum(cluster == j) < thre_ncell) {
                corr[i] <- -10
            }
            print(paste0(" cluster ", j, ";", sum(cluster == 
                j), "cells;", "corr=", corr_, ";diff=", mean(abs(baseline$med_paired - 
                baseline[, i + 3]))))
        }
        i <- which.max(corr)
        j <- unique(cluster)[i]
        if (use_max_cor_cluster) {
            baseline$selected_normal <- apply(norm_count[, cluster == 
                j], 1, median)
        }
        else {
            baseline$selected_normal <- apply(norm_count[, cluster %in% 
                unique(cluster)[which(corr > thre_correlation)]], 
                1, median)
        }
        print(paste0(" selected normal: ", sum(cluster == j), 
            " cells"))
        print(paste0(" max correlation to paired normal: ", corr[i]))
        if (corr[i] < thre_correlation & use_paired) {
            baseline$selected_normal <- baseline$med_paired
            print(" Baseline is calculated using median read count of paired normal sample")
        }
        else {
            print(" Baseline is calculated using median read count of selected normal cells")
        }
    }
    print("Step 4: normalizing against baseline...")
    norm_count_final <- norm_count - baseline$selected_normal
    norm_count_final <- t(norm_count_final)
    norm_count_final <- exp(norm_count_final)
    rownames(norm_count_final) <- colnames(norm_count)[f2]
    q <- quantile(norm_count_final, probs = c(1:100) * 0.01)
    print(paste0("99th percentile: ", q[99]))
    if (is.numeric(cutoff) & cutoff > 1.5) {
        thre1 <- cutoff
    }
    else {
        thre1 <- q[99]
    }
    norm_count_final[norm_count_final > thre1] <- thre1
    colnames(norm_count_final) <- paste0(bin$chr, "_", bin$start, 
        "_", bin$end)[f1]
    print("Step 5: GC correction...")
    norm_count <- t(exp(norm_count))
    colnames(norm_count) <- colnames(norm_count_final)
    norm_re <- list(copy_ratio = norm_count_final, norm_count = norm_count, 
        baseline = exp(baseline$selected_normal))
    tmp <- data.frame(y = norm_re$baseline, gc = bin$gc[f1])
    loess_re <- loess(data = tmp, formula = y ~ gc)
    baseline_gc <- loess_re$fitted
    gc_corrected_cell <- function(y) {
        tmp <- data.frame(y = y, gc = bin$gc[f1])
        loess_re <- loess(data = tmp, formula = y ~ gc)
        y_gc <- loess_re$fitted
        baseline_gc_corrected <- y_gc * norm_re$baseline/baseline_gc
        return(baseline_gc_corrected)
    }
    if (gc_correction) {
        baseline_new <- apply(norm_re$norm_count, 1, gc_corrected_cell)
        norm_re$baseline <- t(baseline_new)
        norm_re$copy_ratio <- norm_re$norm_count/norm_re$baseline
    }
    print("Step 6: saving results...")
    if (!dir.exists(output_dir)) {
        dir.create(output_dir)
    }
    if (output_name != "none") {
        saveRDS(norm_re, file.path(output_dir, output_name))
    }
    return(norm_re)
}

```


```{r}
norm_re = readRDS('/projectnb/paxlab/presh/AtaCNV_output/norm_re.rds')

seg_re <- calculate_CNV_atacnv(norm_count = norm_re$norm_count,
                                baseline = norm_re$baseline,
                                output_dir = "/projectnb/paxlab/presh/AtaCNV_output/",
                                output_name = "CNV_re.rds")
```


```{r}
# Example: Check for common format "chrX:Y-Z"
col_names <- colnames(norm_re$norm_count)

# This regex assumes format like "chr1:1000-2000" or "chrX:1000-2000"
# Adjust regex if your format is different (e.g., "1:1000-2000" or "X_1000_2000")
expected_format_regex <- "^(chr)?([0-9XLY]+):([0-9]+)-([0-9]+)$" 

# Find column names that *do not* match the expected format
malformed_names <- col_names[!grepl(expected_format_regex, col_names)]

if (length(malformed_names) > 0) {
  message("Found malformed column names:")
  print(malformed_names)
} else {
  message("All column names appear to match the expected format according to the regex.")
  # If they all match regex, the issue might be subtler or within str2bin itself for edge cases.
}
```




```{r}
calculate_CNV_atacnv <- function (norm_count, baseline, BICseq = "default", lambda = 5, 
    min_interval = 2, smooth = F, downsample = 0, output_dir = "./", 
    output_name = "CNV_result.rds") 
{
    if (BICseq == "default") {
        if (Sys.info()["sysname"] == "Windows") {
            BICseq <- system.file("BICseq2", "mbic-seq.exe", 
                package = "AtaCNV")
        }
        else if (Sys.info()["sysname"] == "Linux") {
            BICseq <- system.file("BICseq2", "MBICseq", package = "AtaCNV")
            Sys.chmod(BICseq, mode = "0755")
        }
    }
    if (!dir.exists(output_dir)) {
        dir.create(output_dir)
    }
    if (is.vector(baseline)) {
        baseline <- matrix(data = baseline, nrow = nrow(norm_count), 
            ncol = ncol(norm_count), byrow = T)
    }
    bins <- str2bin(colnames(norm_count))
    chr_bkp <- bin2chrbkp(bins)
    print("Running BICseq2 for segmentation...")
    bkp <- c()
    lambda <- lambda
    tmp_dir <- output_dir
    k <- 0
    if (downsample > 0) {
        idx <- sample(1:nrow(norm_count), downsample)
        norm_count_origin <- norm_count
        norm_count <- norm_count[idx, ]
        baseline_origin <- baseline
        baseline <- baseline[idx, ]
    }
    for (i in unique(bins$chr)) {
        print(i)
        # if (sum(bins$chr == i) <= 2) {
        #     next
        # }
        norm_count_ <- norm_count[, bins$chr == i]
        bic_input <- matrix(ncol = 2 * nrow(norm_count_) + 2, 
            nrow = ncol(norm_count_))
        bic_input[, 1:2] <- as.matrix(str2bin(colnames(norm_count_))[, 
            2:3])
        bic_input[, 2 * c(1:nrow(norm_count_)) + 1] <- t(norm_count_)
        bic_input[, 2 * c(1:nrow(norm_count_)) + 2] <- t(baseline[, 
            bins$chr == i])
        s <- c(1:ncol(bic_input))
        s[1:2] <- c("start", "end")
        s[2 * c(1:nrow(norm_count_)) + 1] <- rownames(norm_count_)
        s[2 * c(1:nrow(norm_count_)) + 2] <- paste0(rownames(norm_count_), 
            "_ref")
        colnames(bic_input) <- s
        write.table(x = bic_input, file = paste0(tmp_dir, i, 
            ".bin"), quote = F, sep = "\t", row.names = F, col.names = T, 
            append = F)
        system(paste0(BICseq, " -i ", paste0(tmp_dir, i, ".bin"), 
            " -l ", lambda))
        seg_re <- read.table(paste0(tmp_dir, i, ".bin_seg"), 
            header = T, sep = "\t")
        bkp_ <- cumsum(seg_re$binNum)
        bkp <- merge_bkp(bkp, bkp_ + k, min_interval = 2)
        k <- k + sum(bins$chr == i)
    }
    
    for (i in unique(bins$chr)) {
        file.remove(paste0(tmp_dir, i, ".bin"))
        file.remove(paste0(tmp_dir, i, ".bin_seg"))
    }
    bkp <- merge_bkp(chr_bkp, bkp, min_interval = 2)
    bkp <- sort(bkp)
    print(paste0("number of breakpoints: ", length(bkp)))
    if (downsample > 0) {
        norm_count <- norm_count_origin
        baseline <- baseline_origin
    }
    CNV_re <- matrix(0, nrow = nrow(norm_count), ncol = length(bkp) - 
        1)
    for (i in 1:(length(bkp) - 1)) {
        if (bkp[i] + 1 == bkp[i + 1]) {
            CNV_re[, i] <- norm_count[, bkp[i + 1]]/baseline[, 
                bkp[i + 1]]
            next
        }
        temp <- apply(norm_count[, (bkp[i] + 1):bkp[i + 1]], 
            1, mean)/apply(baseline[, (bkp[i] + 1):bkp[i + 1]], 
            1, mean)
        CNV_re[, i] <- temp
    }
    CNV <- norm_count
    for (i in 1:(length(bkp) - 1)) {
        CNV[, (bkp[i] + 1):bkp[i + 1]] <- matrix(CNV_re[, i], 
            nrow = nrow(CNV), ncol = bkp[i + 1] - bkp[i])
    }
    if (smooth) {
        print("smoothing...")
        CNV <- smooth_CNV(CNV)
    }
    print("Saving results...")
    CNV <- round(CNV, digit = 3)
    result <- list(copy_ratio = CNV, bkp = bkp, CNV_seg = CNV_re)
    saveRDS(result, file = file.path(output_dir, output_name))
    return(result)
}

```

This is giving me too many errors, going to abandon using this method


# CopySCAT 

https://github.com/spcdot/CopyscAT/blob/master/copyscat_tutorial.R
```{r}
# library("devtools")

# Sys.setenv("R_REMOTES_NO_ERRORS_FROM_WARNINGS" = "true")
# install_github("spcdot/copyscat")

#load the package
library(CopyscAT)


setwd('/projectnb/paxlab/presh/Data/copyscat')

#load your favourite genome
library(BSgenome.Hsapiens.UCSC.hg38)
#use it to save references - this will create a genome chrom.sizes file, a cytobands file and a CpG file
#NOTE: updated 21-Oct-2021 with a fallback to the REST API
generateReferences(BSgenome.Hsapiens.UCSC.hg38,genomeText = "hg38",tileWidth = 1e6,outputDir = "/projectnb/paxlab/presh/Data/copyscat/ref")

```



```{r}
##### REGULAR WORKFLOW #####
#initialize the environment (note: this needs to be done with every session in which you run Copy-scAT)
initialiseEnvironment(genomeFile="/projectnb/paxlab/presh/Data/CNV_results/copyscat/ref/hg38_chrom_sizes.tsv",
                      cytobandFile="/projectnb/paxlab/presh/Data/CNV_results/copyscat/ref/hg38_1000000_cytoband_densities_granges.tsv",
                      cpgFile="/projectnb/paxlab/presh/Data/CNV_results/copyscat/ref/hg38_1000000_cpg_densities.tsv",
                      binSize=1e6,
                      minFrags=2e4,
                      cellSuffix=c("-1","-2"),
                      lowerTrim=0.5,
                      upperTrim=0.8)

#for this tutorial we will use the sample data included in scData
#to create your own use the process_fragment_file.py script included in the package and run it on a fragments.tsv.gz file of your choosing

#SET OUTPUT DEFAULT DIRECTORY AND NAME
setOutputFile("/projectnb/paxlab/presh/Data/CNV_results/copyscat","deepseq")

```


Generate table from fragment file RIDDLER 
```{r}
source('/projectnb/paxlab/presh/software/RIDDLER/util/window_file.R')

window_matrix( '/projectnb/paxlab/presh/Data/deepseq.fragments.sort.filtered.bed','/projectnb/paxlab/presh/software/RIDDLER/features/window_1MB.bed', '/projectnb/paxlab/presh/Data/CNV_results/copyscat/counts.txt',frag_min=10000)

scDatasamp = read.table("/projectnb/paxlab/presh/Data/CNV_results/copyscat/counts.txt", header = T, check.names = F)
```




```{r}
#PART 1: INITIAL DATA NORMALIZATION
#step 1 normalize the matrix
#USING SAMPLE DATA FROM PACKAGE
#option: if using your own file replace below with the following
# scData<-readInputTable("myInputFile.tsv")

# Taking counts data from epianeufinder results generated
# counts_summary <- readRDS('/projectnb/paxlab/presh/Data/epianeufinder/atac_results/epiAneufinder_results/count_summary.rds') # for info on ranges

# count = assay(counts_summary)
# rownames(count) = paste(counts_summary@rowRanges$wSeq, counts_summary@rowRanges$wStart, counts_summary@rowRanges$wEnd, sep = '_')



# scDataSamp <- t(count)

scData<-read.table('/projectnb/paxlab/presh/Data/deepseq.fragments.sort.filtered.bed')

scData_k_norm <- normalizeMatrixN(scData,logNorm = FALSE,maxZero=2000,imputeZeros = FALSE,blacklistProp = 0.8,blacklistCutoff=125,dividingFactor=1,upperFilterQuantile = 0.95)
#when using your own data, please make sure you don't have any excess / alt chromosomes
#collapse into chromosome arm level
summaryFunction<-cutAverage

scData_collapse<-collapseChrom3N(scData_k_norm,summaryFunction=cutAverage,binExpand = 1,minimumChromValue = 100,logTrans = FALSE,tssEnrich = 1,logBase=2,minCPG=300,powVal=0.73) 

#apply additional filters
scData_collapse<-filterCells(scData_collapse,minimumSegments = 40,minDensity = 0.1)

#show unscaled chromosome list
graphCNVDistribution(scData_collapse,outputSuffix = "test_violinsn2")

#compute centers
median_iqr <- computeCenters(scData_collapse,summaryFunction=summaryFunction)

```



# RIDDLER 

Date: 2025-11-28 

Generate table from fragment file RIDDLER 
```{r}
source('/projectnb/paxlab/presh/software/RIDDLER/util/window_file.R')

window_matrix( '/projectnb/paxlab/presh/Data/deepseq.fragments.sort.filtered.bed','/projectnb/paxlab/presh/software/RIDDLER/features/window_1MB.bed', '/projectnb/paxlab/presh/Data/CNV_results/copyscat/counts.txt',frag_min=10000)

scDatasamp = read.table("/projectnb/paxlab/presh/Data/CNV_results/copyscat/counts.txt", header = T, check.names = F)
```


```{r}
source("/projectnb/paxlab/presh/software/RIDDLER/riddler_fit.R")

setwd('/projectnb/paxlab/presh/software/RIDDLER/')
mat_file = "/projectnb/paxlab/presh/Data/CNV_results/copyscat/counts.txt"

rid_obj = riddler_fit(mat_file,feat_file="features/features.1MB.csv",peak_file="features/tissue_peaks.1MB.bed",cov_min=10000,peak_corr_filter=.5)
```


```{r}
setwd('/projectnb/paxlab/presh/software/RIDDLER/')
source("copy_call.R")

rid_obj = copy_call(rid_obj,w_size=5,prob_thresh=.025,mult=c(0,.5,1,1.5,2,3))

saveRDS(rid_obj, '/projectnb/paxlab/presh/Data/CNV_results/riddler/rid_obj.rds')
```

```{r}
setwd('/projectnb/paxlab/presh/software/RIDDLER/')
source("cnv_heatmap.R") 

rid_obj = readRDS( '/projectnb/paxlab/presh/Data/CNV_results/riddler/rid_obj.rds')

heat_plot = cnv_heatmap(rid_obj,heat_title="RIDDLER CNVs",ref_bar=NULL,ref_title=NULL,cluster=TRUE)

ggsave(plot=heat_plot,"/projectnb/paxlab/presh/Data/CNV_results/riddler/heatmap.png",device="png",height=5,width=8)
```






