---
title: "6_compare_atac_xenium"
editor_options: 
  chunk_output_type: console
---

Date: 23 October 2025
Comparison between Atac seq and Xenium data clusters 

Let's see if the data more or less corresponds between them 

# Setting up libraries
```{r}
.libPaths('/projectnb/paxlab/presh/giotto_libs')
.libPaths()

library(Giotto)
library(ggplot2)

# Ensure Giotto can access the python env
set_giotto_python_path(python_path = '/usr3/graduate/preshita/.conda/envs/giotto_env/bin/python', verbose = NULL, initialize = TRUE)

```


# Xenium obj
## Data load in 
```{r}

data_path = '/projectnb/paxlab/DATA/DriesSpatial/Xenium/output-XETG00253__0048833__488B__20241217__182301'


results_folder <- '/projectnb/paxlab/presh/giotto_output'

instrs <- createGiottoInstructions(
  show_plot = T,
  return_plot = FALSE,
  save_plot = FALSE,
  save_dir = results_folder
)

data_path <- '/projectnb/paxlab/DATA/DriesSpatial/Xenium/output-XETG00253__0048833__488B__20241217__182301'

# x <- importXenium(data_path)
# force(x)
# 
# xenium488B  <- giotto(instructions = instrs) |>
#   setGiotto(x$load_polys()) |>
#   setGiotto(x$load_expression('/projectnb/paxlab/DATA/DriesSpatial/Xenium/output-XETG00253__0048833__488B__20241217__182301/cell_feature_matrix/')) |>
#   setGiotto(x$load_cellmeta()) |>
#   addSpatialCentroidLocations()
# 
# force(xenium488B)
# 
# plot(xenium488B@spatial_locs$cell$raw)
# 
# 
# 
# xenium488B = filterGiotto(xenium488B, feat_det_in_min_cells = 100, 
#                   min_det_feats_per_cell = 20, 
#                   expression_threshold = 1)
# 
# norm_name <- 'normalized'
# xenium488B = normalizeGiotto(xenium488B)
```


morphology_focus/
├── ch0000_dapi.ome.tif (DAPI)
├── ch0001_atp1a1_e-cadherin_cd45.ome.tif (ATP1A1/E-Cadherin/CD45)
├── ch0002_18s.ome.tif (18S rRNA)
└── ch0003_alphasma_vimentin.ome.tif (alphaSMA/Vimentin)

```{r}
# Making a named list of images
img_list = list('dapi' = "/projectnb/paxlab/presh/tif_exports/morphology_focus_0000.tif", 
                'img_1' = "/projectnb/paxlab/presh/tif_exports/morphology_focus_0001.tif",
                'img_2' = "/projectnb/paxlab/presh/tif_exports/morphology_focus_0002.tif",
                'img_3' = "/projectnb/paxlab/presh/tif_exports/morphology_focus_0003.tif" )


feat_types <- c(
    "rna",
    "UnassignedCodeword", 
    "NegControlCodeword",  
    "NegControlProbe"
)
split_keywords = list(
    c("BLANK"),
    c("NegControlCodeword"),
    c("NegControlProbe", "antisense")
)


xenium488B = createGiottoXeniumObject(
  data_path,
  transcript_path = NULL,
  bounds_path = list(cell = "cell", nucleus = "nucleus"),
  gene_panel_json_path = NULL,
  expression_path = '/projectnb/paxlab/DATA/DriesSpatial/Xenium/output-XETG00253__0048833__488B__20241217__182301/cell_feature_matrix/',
  cell_metadata_path = NULL,
  feat_type = feat_types,
  split_keyword = split_keywords,
  qv_threshold = 20,
  load_images = img_list,
  load_aligned_images = NULL,
  load_transcripts = F,
  load_expression = T,
  load_cellmeta = T,
  instructions = instrs,
  verbose = NULL
)


force(xenium488B)
plot(xenium488B@spatial_locs$cell$raw)

# let's crop the object 

spatPlot2D(xenium488B,
    # scattermore speeds up plotting
    plot_method = "scattermore",
    point_shape = "no_border",
    point_size = 1
)


# cropping the xenium object roughly 
xenium_crop <- subsetGiottoLocs(xenium488B, x_min = 5000, x_max = 10600)


```

Loading transcripts into the object takes forever, not needed
Also we don't aggregate the data as it needs feat_type which comes from loading transcripts 

GOING TO ONLY WORK WITH XENIUM CROP IMAGE FOR NOW!


## normalize and pca 
```{r}
xenium_crop <- xenium_crop |>
    filterGiotto(,
        spat_unit = 'cell',
        expression_threshold = 1,
        feat_det_in_min_cells = 3,
        min_det_feats_per_cell = 5
    ) |>
    normalizeGiotto() |> 
    addStatistics() |>
    runPCA(feats_to_use = NULL) # don't use HVFs since there are too few features*
```


```{r}
# Visualize Screeplot and PCA
screePlot(xenium_crop, ncp = 20)
```


```{r}
plotPCA(xenium_crop, point_size = 0.1)
```


## leiden clustering 
```{r}

xenium_crop = createNearestNetwork(gobject = xenium_crop,
                                        dimensions_to_use = 1:10,
                                        k = 10, # number of neighbors
                                        name = 'nn_network'
                                        )

xenium_crop = doLeidenCluster(gobject = xenium_crop,
                                   network_name = 'nn_network', # use the network created in step 3
                                   name = 'leiden_clus', # choose a name for the cluster results
                                   resolution = 0.05 # adjust resolution as needed
                                   )
```

## umap 
```{r}
xenium_crop <- runUMAP(gobject = xenium_crop,
                   dimensions_to_use = 1:10, # adjust based on your PCA results
                   n_components = 2,       # set to 3 for 3D, or 2 for 2D
                   n_threads = 4)

# Plot the UMAP colored by the Leiden clusters
plotUMAP_2D(gobject = xenium_crop,
            cell_color = 'leiden_clus',
            point_size = 1
            )
```


## visualize clusters spatially 

```{r}
spatPlot2D(gobject = xenium_crop,
                          spat_unit = 'cell',
                          cell_color = 'leiden_clus',
                          point_size = 0.025,
                          point_shape = 'no_border',
                          # color_as_factor = F,
                          show_legend = TRUE,point_border_stroke = 0,
                          background_color = 'black',
                          save_param = list(
                            save_name = 'sg_spat_leiden',
                            base_width = 15,
                            base_height = 15))

```



```{r}
spatFeatPlot2D(xenium_crop, 
              expression_values = "scaled", 
              feats = c('TERT', 'FOXA1', 'CD19', 'CPA3', 'KLRD1'), cow_n_col = 2, 
              point_size = 0.2)


# Breast cancer genes
spatFeatPlot2D(xenium_crop, 
              expression_values = "normalized",  cell_color_gradient = c('white', 'red'),
              feats = c('ESR1', 'ERBB2', 'PGR', 'MKI67'), gradient_limits = c(0,3), 
              point_size = 0.1)


```

ERBB2 - also HER2



<!-- ```{r} -->
<!-- # example plot for polygons -->
<!-- spatInSituPlotPoints(xenium488B, -->
<!--     xlim = c(5500, 6000), -->
<!--     ylim = c(-2000, -1000), -->
<!--     show_image = TRUE, -->
<!--     image_name = "dapi", -->
<!--     polygon_line_size = 0.1, -->
<!--     polygon_color = "#BB0000", -->
<!--     polygon_alpha = 0.2 -->
<!-- ) -->
<!-- ``` -->



# Atac object

Loading in the imputed object 
```{r}
# atac.obj <- readRDS('/projectnb/paxlab/presh/saveAtacgiotto.rds')
atac.imp.obj <- readRDS('/projectnb/paxlab/presh/saveAtacgiotto_imputed.rds')

sg <- atac.imp.obj$Deepseq

# we should filter out more cells since we know that the borders are not great 
int.cells.atac = read.csv('/projectnb/paxlab/presh/Data/int_cells_trim_atac.csv', header = T, row.names = 1)
int.cells.atac =int.cells.atac$x

sg2 = subsetGiotto(
  sg,
  spat_unit = ":all:", feat_type = ":all:", 
  cell_ids = int.cells.atac
)

# sg2 = filterGiotto(sg2, feat_det_in_min_cells = 100,
#                   spat_unit = 'cell',
#                   feat_type = 'atac',
#                   expression_values = 'GeneScoreMatrix',
#                   min_det_feats_per_cell = 20, 
#                   expression_threshold = 1)
# 
# 
# norm_name <- 'normalized'

activeSpatUnit(sg2) <- 'cell'
activeFeatType(sg2) <- 'atac'


sg2@expression[["cell"]][["atac"]][["normalized"]] = sg2@expression[["cell"]][["atac"]][["GeneScoreMatrix"]]
sg2@expression[["cell"]][["atac"]][["scaled"]] = scale(sg2@expression[["cell"]][["atac"]][["normalized"]]@exprMat)


# sg2 = normalizeGiotto(sg2, norm_methods = 'standard', name = norm_name, expression_values = 'GeneScoreMatrix')

# Statistics
# sg2 = addStatistics(sg2, expression_values = norm_name)


# test <- spatPlot2D(gobject = sg2,
#                    show_image = F,
#                    point_alpha = 0.7,
#                    point_size = 0.6,
#                    background_color = 'black',
#                    cell_color = 'nr_feats',
#                    color_as_factor = F,
#                    save_param = list(
#                      save_name = 'nr_feats',
#                      base_width = 15,
#                      base_height = 15))
# 
# test

```


```{r}
spatFeatPlot2D(sg2, expression_values = "scaled",
              feats = c('TERT', 'FOXA1', 'CD19', 'KLRD1','SOX17','MS4A1','HOXD9','KLRC1', 'TPSAB1'))

```

Breast cancer genes
```{r}
# scaled expr
spatFeatPlot2D(sg2, expression_values = "scaled",
              feats = c('ESR1', 'MKI67', 'EGFR', 'ERBB2', 'PGR'))

# just normalized expression 
spatFeatPlot2D(sg2, expression_values = "normalized", 
              feats = c('ESR1'),  cell_color_gradient = c('white', 'red'), gradient_limits = c(0,3), 
              point_size = 1)

```


I'm not able to force my limits weirdly 


<!-- ## umap -->
<!-- ```{r} -->


<!-- sg2 = runPCA(feats_to_use = NULL) # don't use HVFs since there are too few features* -->

<!-- screePlot(sg2) -->

<!-- # Run and Plot tSNE and UMAP -->
<!-- sg2 = runUMAP(sg2, -->
<!--              expression_values = 'normalized', -->
<!--              dimensions_to_use = 1:13, -->
<!--              spat_unit = 'cell') -->

<!-- plotUMAP(sg2, -->
<!--          point_size = 0.01, -->
<!--          save_param = list( -->
<!--            save_name = 'sg_UMAP')) -->

<!-- # Clustering and UMAP cluster visualization -->
<!-- sg2 = createNearestNetwork(sg2, -->
<!--                           dimensions_to_use = 1:13, -->
<!--                           expression_values = norm_name, -->
<!--                           k = 25, -->
<!--                           spat_unit = 'cell') -->
<!-- sg2 = doLeidenClusterIgraph(sg2, -->
<!--                      resolution = 0.4, -->
<!--                      n_iterations = 10, -->
<!--                      spat_unit = 'cell') -->

<!-- # Plot Leiden clusters onto UMAP -->
<!-- plotUMAP(gobject = sg2, -->
<!--          spat_unit = 'cell', -->
<!--          # select_cells = healthy_cancer, -->
<!--          cell_color = 'leiden_clus', -->
<!--          show_legend = TRUE, -->
<!--          point_size = 0.01, -->
<!--          point_shape = 'no_border', -->
<!--          save_param = list(save_name = 'sg_umap_leiden')) -->

<!-- my_spatPlot <- spatPlot2D(gobject = sg2, -->
<!--                           spat_unit = 'cell', -->
<!--                           cell_color = 'leiden_clus', -->
<!--                           point_size = 0.025, -->
<!--                           point_shape = 'no_border', -->
<!--                           # color_as_factor = F, -->
<!--                           show_legend = TRUE,point_border_stroke = 0, -->
<!--                           background_color = 'black', -->
<!--                           save_param = list( -->
<!--                             save_name = 'sg_spat_leiden', -->
<!--                             base_width = 15, -->
<!--                             base_height = 15)) -->
<!-- ``` -->

## add metadata
```{r}
archr.meta = read.table('/projectnb/paxlab/presh/Data/archr_metadata.tsv')
archr.meta$cell_ID =gsub('-1','' ,gsub('Deepseq#', '', rownames(archr.meta)))

sg2 <- addCellMetadata(gobject = sg2,
                               new_metadata = archr.meta, by_column = T, 
                               column_cell_ID = 'cell_ID')
```




# Similarity between clusters

Date: 2025-11-10

Jeff reported that some genes are anti-correlated in Xenium vs Atac seq data

Let's look at cluster similarity gene by correlation maybe?

```{r}

avg_expr_xenium = calculateMetaTable(
    gobject = xenium_crop,
    expression_values = "normalized", # or "scaled", "custom", etc.
    metadata_cols = "leiden_clus"
)


```

I don't care about cluster 5 xenium 

```{r}
avg_expr_atac = calculateMetaTable(
    gobject = sg2,
    expression_values = "normalized", # or "scaled", "custom", etc.
    metadata_cols = "Clusters_tile"
)
avg_expr_atac = na.omit(avg_expr_atac)

# remove genes not in xenium 
avg_expr_atac_filt = avg_expr_atac[avg_expr_atac$variable %in% unique(avg_expr_xenium$variable), ]
avg_expr_atac_filt$uniq_ID <- as.numeric(gsub('C','',avg_expr_atac_filt$uniq_ID))
```

```{r}
library(tidyr)
library(dplyr)
library(tibble)

df1 <- avg_expr_atac_filt
df2 <- avg_expr_xenium

# Make matrices (clusters × variables)
atac_mat <- df1 %>% 
  pivot_wider(names_from = variable, values_from = value) %>% 
  column_to_rownames("uniq_ID")

xenium_mat <- df2 %>% 
  pivot_wider(names_from = variable, values_from = value) %>% 
  column_to_rownames("uniq_ID")

atac_mat <- atac_mat[, -1]
xenium_mat <- xenium_mat[, -1]
```


```{r}
library(tidyr)
#1. Compute Expression Profiles for Each Cluster
atac_cluster_expr <- df1 %>% pivot_wider(names_from = variable, values_from = value)
xenium_cluster_expr <- df2 %>% pivot_wider(names_from = variable, values_from = value)

#2. Normalize Each Cluster's Profile
# Make sure all columns except uniq_ID are numeric
atac_norm <- atac_cluster_expr %>%
  select(-uniq_ID, -Clusters_tile) %>% 
  mutate(across(everything(), as.numeric)) %>%
  as.matrix() %>%
  t() %>%
  scale(center=TRUE, scale=TRUE) %>%
  t()

xenium_norm <- xenium_cluster_expr %>%
  select(-uniq_ID, -leiden_clus) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix() %>%
  t() %>%
  scale(center=TRUE, scale=TRUE) %>%
  t()


shared_genes <- intersect(colnames(atac_norm), colnames(xenium_norm))
atac_sub <- atac_norm[, shared_genes]
xenium_sub <- xenium_norm[, shared_genes]
#3. Calculate Similarity Matrix between Clusters Across Assays
similarity_matrix <- cor(t(atac_sub), t(xenium_sub), use="pairwise.complete.obs")
# Rows: ATAC clusters, Columns: Xenium clusters


```

```{r}
max_match_atac <- apply(similarity_matrix, 1, function(x) colnames(similarity_matrix)[which.max(x)])
max_similarity_atac <- apply(similarity_matrix, 1, max)

# As a dataframe:
cluster_matching <- data.frame(
  ATAC_cluster = rownames(similarity_matrix),
  Best_Xenium_cluster = max_match_atac,
  similarity = max_similarity_atac
)
```

```{r}
library(pheatmap)
pheatmap(similarity_matrix, display_numbers = TRUE, cluster_rows = FALSE, 
         cluster_cols = FALSE,
         main = "Cluster similarity: ATAC (Y axis) vs Xenium (X axis)")
```





```{r}
compare_binary <- function(var) {
  a <- df1 %>% filter(variable == var) %>% select(uniq_ID, atac_value = value)
  x <- df2 %>% filter(variable == var) %>% select(uniq_ID, xenium_value = value)
  joined <- inner_join(a, x, by = "uniq_ID")
  joined <- joined %>%
    mutate(atac_bin = atac_value > median(atac_value),
           xenium_bin = xenium_value > median(xenium_value),
           agreement = case_when(
             atac_bin & xenium_bin ~ "both_high",
             !atac_bin & !xenium_bin ~ "both_low",
             atac_bin & !xenium_bin ~ "ATAC_high_Xenium_low",
             !atac_bin & xenium_bin ~ "ATAC_low_Xenium_high"
           ))
  return(joined)
}
agreement <- lapply(common_vars, compare_binary)
```










